The adaptive landscapes of three global Escherichia coli transcriptional regulators

Version 1

Abstract

Abstract

The evolution of new gene regulation is an important source of evolutionary adaptations and innovations, especially when organisms encounter new environments. At its heart is the process by which new and strong DNA binding sites of transcription factors (TFs) originate in evolution. Here we study the potential of Darwinian evolution to create strong binding sites for the three Escherichia coli global transcription factors CRP, Fis, and IHF. Using a massively parallel reporter assay, we measure for each TF the ability of more than 30,000 binding sites to regulate gene expression. We use the resulting data to map the adaptive landscape of each TF. We find that all three landscapes are rugged, epistatic, and harbor multiple peaks. The highest peaks are widely scattered throughout the landscape, indicating that strong regulation can be achieved by very different binding sites. Landscape ruggedness does not prevent the evolution of strong regulation, because more than 10% of evolving populations can attain one of the highest peaks. Adaptive evolution starting from the same DNA sequence can attain more than one high peak, and some high peaks are more likely to be reached than others. Our experiments show that de novo adaptive evolution of new gene regulation is feasible. It is also subject to a blend of chance, historical contingency, and evolutionary biases that favor some peaks and evolutionary paths over others.

Introduction

Transcriptional regulation controls the ability of RNA polymerase to initiate the transcription of a gene into mRNA1. It is crucial in the life of all organisms, orchestrating the expression of genes in response to environmental cues and cellular states2,3. Transcriptional regulation is mediated by transcription factors (TFs), proteins that bind DNA near a gene at short DNA words known as transcription factor binding sites (TFBSs). The binding of a TF to its binding site on DNA can either hinder or facilitate transcription initiation by RNA polymerase1,4,5. The stronger the binding of a TF to its TFBS is, the more strongly the TF can regulate a nearby gene6–9. In bacteria, TFs operate within a hierarchically organized gene regulatory network. At the bottom of this hierarchy are local TFs that regulate the expression of one or few genes and modulate specific biological processes 10–12. At the top are global TFs that may regulate hundreds of genes and control many cellular functions 11,13–15. Consistent with their broader role, global regulators are typically more highly expressed and bind to a broader range of TFBSs than local regulators12.

Both TFs and their TFBSs evolve, but TFBSs evolve more rapidly. The reason is that a mutation in a TF can affect the expression of many genes, whereas a mutation in a TFBS may affect only one gene and is less likely to be deleterious 13,16–18. A special case of TFBS evolution is the evolution of a strong TFBS from a DNA word with weak or no regulatory activity. Such de novo evolution of TFBSs can create new regulatory interactions and change the structure of gene regulatory networks19–22. Unfortunately, we know little about the ability of Darwinian evolution to create TFBSs de novo. A strong TFBS may have to arise from a non-binding site through an evolutionary path of multiple mutational steps. Darwinian evolution can favor this process only if strong binding is adaptive and if each mutational step in a path increases binding strength, i.e., if the path is accessible to Darwinian evolution. To find whether such paths exist may require the analysis of multiple paths. This is challenging because sequence space contains an astronomical number of potential TFBSs for any one TF. The number of evolutionary paths to strong transcriptional regulation is thus also astronomical. For each such path, the strength of each TFBS along the path has to be measured experimentally23–26.

Building on our previous work on a local TF27, we take a step to address this challenge for three global regulators in Escherichia coli. Specifically, we perform three independent and massively parallel experiments23 using a plasmid-based expression system for each TF. Our synthetic biology approach allows us to explore the regulatory effects of tens of thousands of TFBS variants in a high-throughput manner. We quantify how strongly each of more than 30,000 binding sites for each TF can regulate the expression of a nearby reporter gene.

The first TF we study is the cAMP receptor protein (CRP). In the absence of glucose, CRP modulates the expression of genes mostly involved in carbon metabolism. It allows E. coli to efficiently switch between sources of carbon and energy 28–31. The second TF is the factor for inversion stimulation (Fis), which helps to regulate the expression of genes involved in growth phase transitions. It also modulates the supercoiling of DNA32–34 and influences DNA replication, recombination, and repair32–34. The third factor is the integration host factor (IHF). IHF regulates genes involved in stress responses and stationary phase survival35,36. Similar to Fis, it is involved in DNA compaction, replication, and recombination, but also in the assembly of complex nucleoprotein structures35,36. We chose these factors because they are the most global regulators in E. coli, and they are diverse, belonging to different protein families.

We use the regulatory data for each TF’s binding sites to map a genotype-phenotype landscape for these binding sites 37,38. Such a landscape is a conceptual analog of a physical landscape, where each location corresponds to one genotype in sequence space, and the elevation at this location corresponds to a phenotype – the ability to regulate gene expression 6,27,39. One of our two main aims is to provide the first characterization of such landscapes for global bacterial regulators, and to find out whether these landscapes are different or similar.

When strong regulation of a gene is adaptive, a regulatory landscape becomes an adaptive landscape or fitness landscapes 38,40–42. Our second aim is to understand how populations would evolve on each of our three landscapes when they are viewed as fitness landscapes. Specifically, we study how evolving populations would traverse each landscape through individual mutational steps that change a TFBS its ability to regulate a gene via its cognate TF. A peak in such a landscape is a TFBS conveying stronger regulation than all neighboring TFBSs in sequence space. If such a landscape is rugged (has multiple peaks), natural selection alone may not enable a population to discover the highest peaks, i.e., the strongest TFBSs. The reason is that the peaks may be separated by valleys of low regulation strength that cannot be traversed by natural selection alone 43–45 38. In other words, high peaks may not be easily accessible through Darwinian evolution – they may be reachable by few or no evolutionary paths of single DNA mutations where each mutational step increases TFBS strength6. One factor that can reduce peak accessibility is epistasis – non-additive effects of two or more mutations on a phenotype. In addition, epistasis can reduce the predictability of evolutionary trajectories towards a peak 37,46.

To accomplish both aims, we first studied the topography of the three landscapes, and then simulated the dynamics of evolving populations on them. All three landscapes are highly rugged. They contain more than 2,000 peaks that are scattered through genotype space and are also rife with epistatic interactions. Despite these features, evolving populations can reach the strongest TFBSs in all three landscapes. Which of several high peak is reached is contingent on chance events during adaptive evolution.

Results

Landscape mapping

We constructed a plasmid system (Supplementary Table S1, Supplementary Figures S1-S2 and Supplementary Methods 2-3) to inducibly express each of the three global TFs CRP, Fis, and IHF in E. coli. We expressed each factor in a host strain in which its encoding chromosomal gene had been knocked out (Δcrp, Δfis and Δihf, see Supplementary Table S2). This ensures that each TF is expressed only upon induction. Expression of each TF can be induced with anhydrotetracycline (atc), which allowed us to control TF expression levels and measure how the binding of the TF to DNA can influence transcriptional regulation by measuring the expression of a green fluorescent reporter gene (gfp) (Figure 1a). In our plasmid system, stronger binding of a TF to DNA leads to lower GFP expression. The reason is that the TF’s binding site is located directly upstream of the gfp gene, and TF-DNA binding thus downregulates GFP expression by interfering with gfp transcription. Our three plasmid-strain pairs attained similar cell densities during late exponential or early stationary phase (Supplementary Figure S3) where we performed all our measurements.

Mapping TFBS genotype-phenotype landscapes.

a-c) Sort-Seq procedure. We utilized a plasmid-based fluorescence reporter system followed by sort-seq to map TFBS genotype-phenotype landscapes. a. Library generation. For generating our E. coli libraries, we cloned TFBS sequence variants (28 = 65,536), represented as black circles, into our plasmid, between a sigma70 constitutive promoter (shown as a black right-facing arrow) and a gfp gene (shown as a green right-facing arrow), to measure transcriptional repression through fluorescence intensity. When a TF binds to the TFBS, it blocks RNA polymerase activity through steric hindrance, reducing gfp transcription. Thus, lower fluorescence levels (light green) represent stronger TF-TFBS binding. The TFBS variants produce a range of regulation strengths resulting, in variable green fluorescence intensities in bacterial cells (green-colored rectangles). b. Sorting procedure. We sorted libraries into expression bins based on fluorescence intensity using a fluorescence-activated cell sorting (FACS). c. Sequencing. We then sequenced the TFBSs from each bin and mapped their sequence (genotype) to their respective regulation strengths, depicted through a color gradient from dark green (low-affinity TFBS, high fluorescence) to light green (high-affinity TFBS, low fluorescence). d) Genotype-phenotype mapping. To construct a regulatory landscape, we connected TFBS genotypes (colored circles) that differed by a single nucleotide via edges, thereby establishing an interconnected genotype network. Each genotype conveys a regulatory phenotype (strength of regulation, heatmap colors), which can be viewed as the elevation dimension (z-axis) in a landscape.

For each plasmid expressing one of our three TFs, we cloned an entire library of TFBSs upstream of the gfp reporter. Specifically, in this library, we randomized each of the eight most important (information-rich) base pair positions within the TFBS (see Supplementary Methods 4 and Supplementary Table S3). We obtained these most information-rich positions for each TF from the alignment of known TFBSs in the RegulonDB database47. Each library comprised 48=65,536 unique sequences. We constructed three replicates of each library (Supplementary Methods 4 and 5).

We then mapped these TFBS genotypes to their respective regulatory phenotypes using a well-established technique known as sort-seq23,39,48–50 (Figure 1b), which combines fluorescence-activated cell sorting (FACS) with high-throughput sequencing (Figure 1c). In sort-seq, one first sorts cells into multiple fluorescence “bins” depending on the level of GFP expression. A cell’s GFP fluorescence serves as a proxy for GFP expression levels51 and quantifies how strongly the TFBS library member in this cell can regulate GFP expression (Methods, Supplementary Methods 5).

The results of our sort-seq experiments are three maps – one for each transcription factor – from each of more than 30,000 genotypes (TFBS variants) to regulatory phenotypes (regulation strength). One can view each map as a genotype-phenotype landscape that we also call a regulatory landscape8,9,27,39,52–56 (Figure 1d). Whenever strong regulation entails high fitness, such a landscape is also an adaptive landscape or fitness landscape6,38. For the purpose of analyzing the landscape quantitatively, we represent it as a network of genotypes (TFBS variants). Each node in this network corresponds to a genotype. Edges link neighboring genotypes, which differ in a single nucleotide (Figure 1d).

Landscapes exhibit diverse regulation strengths and distribution breadths

To evaluate the ability of our library to regulate gene expression, we first measured the distribution of GFP fluorescence intensities across the bins in two conditions, i.e., in the presence or absence of the TF (with or without the atc inducer, Supplementary Figures S4-S6). In the presence of the TF, GFP expression was lower on average and showed a broader distribution than in the absence of the TF (Supplementary Figures S4-S6). This indicates that the TFBSs in each library can indeed downregulate GFP expression, but some library variants convey stronger regulation than others, hence the broader fluorescence distribution.

We then pooled barcoded DNA sequences extracted from cells in each bin and biological replicate. We sequenced at least 250 unique TFBS genotypes from each bin, with an average of 3992.3±4293.9 unique sequences per bin for the three TFs (as detailed in Methods, Supplementary Methods 7, and Supplementary Figures S4-S6). The resulting sequences covered 95%, 90%, and 93% of the total library sizes (N = 65,536 genotypes) for CRP, Fis and IHF, respectively. To ensure the reliability of our data, we excluded sequences with low read coverage and sequences that were not present in all triplicates (Methods, Supplementary Methods 7.3). This quality filtering step resulted in library sizes of 31,975 genotypes for CRP (49% of all 48 genotypes), 43,222 genotypes for Fis (66%), and 41,325 genotypes for IHF (63%). The correlation in read counts for each variant across replicates was very high for this quality-filtered data (Pearson’s R= 0.98-0.99) (Supplementary Figures S7-S9). We used this data for all further analyses.

The fluorescence and sequence data from each bin allowed us to quantify the variant’s ability to regulate gene expression for each TFBS library variant. We refer to the resulting metric as the regulation strength S conveyed by the variant (Methods, Supplementary Methods 7.2, Figure 1d). It ranges between S=0 (no regulation) to S=1 (strongest regulation among all variants in the library). Although our experiments directly quantify expression regulation, and not the affinity or binding strength of a TF to a TFBS variant, we also use binding strength as a proxy for regulation strength, because TF-DNA binding is necessary for regulation1,4,11,57–59. To each of our three TFs, we also assigned a reference TFBS that is naturally occurring and conveys strong regulation by the TF, as proven by previous experimental work31,60–62. We refer to this TFBS as the wild-type (WT 60,63 WT 61,64 and WT 62,65, Supplementary Table S4). We refer to TFBSs with regulation strengths below and above the wild-type (WTCRP: SWT=0.71, WTFis: SWT=0.97, WTIHF: SWT=0.95) as weak and strong, respectively.

We observed a broad range of regulation strengths S for each TF landscape, with varying spreads. The CRP landscape exhibited the lowest average regulation strength and a narrower distribution of S compared to Fis and IHF, which had similar distributions (mean±s.d., 0.37±0.1 for CRP, 0.57±0.13 for Fis, and 0.52±0.14 for IHF; see Figure 2a, Supplementary Figure S10). Next, we analyzed the strongly regulating TFBSs to identify nucleotides that may be particularly frequent and thus, potentially important for strong regulation (Figure 2b). We discovered a moderate association between the most frequent nucleotides (highlighted in yellow in Figure 2b) and the most informative nucleotides from the available position weight matrices (PWMs) for these TFs66. (Pearson correlation coefficients: R=0.51, R=0.43, R=0.47 for Fis, CRP and IHF, respectively, with p-values smaller than 10-16, rejecting the null hypothesis of an absence of association). This observation suggests that our logos capture different information compared to available PWMs. This is expected, because our approach allows us to filter sequences by regulatory strength thresholds to construct our matrices, unlike the traditional method of aligning genomic TFBSs without considering their regulation strengths67,68.

a. Genotypes in each landscape vary broadly in their regulation strength. The violin plots show the distribution of regulation strengths S (vertical axis) for each TF landscape (horizontal axis). The width of a plot at a given value of S represents the frequency of TFBSs at this value. The vertical length of the box in each violin plot covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR. The white circle shows the regulation strength of the wild type for each landscape (CRP: 0.71, Fis: 0.97, IHF: 0.95). The landscape sizes are as follows. CRP: N=31,975 TFBS variants; Fis: N= 43,222; IHF: N= 41,325. b. Peak genotypes are stronger regulators than non-peak genotypes. Dual violin plots show the distribution of regulation strength (vertical axis) for the three TF landscapes (horizontal axis), stratified by non-peak genotypes (grey, CRP, N=29,821, Fis, N= 40,910, IHF, N= 38,872), and peak genotypes (red, CRP, N= 2,154, Fis, N= 2,312, IHF, N= 2,453). The black tick-mark in each plot indicates the mean regulation strength of both non-peak and peak genotypes taken together (mean ± standard deviation, CRP: 0.37 ±0.1, Fis: 0.57±0.13, IHF: 0.52±0.14). The white circle on each plot marks the regulation strength of the wild-type (CRP: 0.71, Fis: 0.97, IHF: 0.95). c. Sequence logos and nucleotide frequency matrices for strong CRP, Fis and IHF binding sites. Each sequence logo 67,68 is based on an alignment of TFBSs with greater regulation strength than the wild-type for each of the three TFs IHF, Fis, and CRP. Each logo also shows the non-varying position (grey) of the TFBS genotype from which our libraries were created. In each logo, the height of each letter at each TFBS position indicates the information content at that nucleotide position – the taller the letter, the more frequent the nucleotide is in strongly regulating TFBSs 67,68. Similar information is conveyed by the frequency matrices displayed as heat maps below each logo. They represent the variability of each nucleotide at each position (horizontal axis) through a color gradient (see color legend). Tall letters in the sequence logo and yellow letters in the frequency matrix indicate frequent, and thus likely important nucleotides for strong regulation.

To further validate the data from our sort-seq experiments, we isolated cells harboring 10 different TFBS variants from each of 13 bins of each library (i.e., 10×13=130 variants per library), and determined their regulation strength more directly by quantifying GFP expression with a microplate reader (Methods and Supplementary Figure S11). This comparison validates the sort seq approach by revealing a strong association of the two independent quantifications of regulation strength (Pearson’s R=-0.81 for CRP, R=-0.74 for Fis, and R=-0.73 for IHF). (Methods and Supplementary Figure S11).

All three regulatory landscapes are highly rugged

The study of our landscapes in a network framework (Figure 1d) can help to quantify different aspects of landscape topography (Supplementary Table S5). One of them is the ruggedness of each landscape. It can be quantified by the number of peaks69–72. In the network framework, a peak is a TFBS whose neighbors are all weaker regulators (with lower S) than the TFBS itself. We find that all three landscapes are highly rugged, with 2,154, 2,312, and 2,453 peaks for the CRP, Fis, and IHF landscapes, respectively (Figure 2c, Supplementary Figure S12, Supplementary Table S5). Not surprisingly, peak genotypes generally are stronger regulators than non-peak genotypes (Figure 2c). Only a small fraction of peak genotypes regulate expression more strongly than the wild-type (Figure 2c; 61 for CRP, 172 for Fis, and 199 for IHF). We refer to such peaks as high peaks and distinguish them from low peaks (S<SWT).

We compared the ruggedness of our landscapes with that of a well-established theoretical model of uncorrelated random landscapes, in which each sequence is assigned a fitness at random from the same fitness distribution, and neighbors have uncorrelated fitness values72–74 Such landscapes are maximally rugged 72–74. We created 100 uncorrelated random landscapes for each of our three TF landscapes by randomly shuffling the measured regulation strengths among all genotypes (Supplementary Methods 7.6). The number of peaks in our TF landscapes lies within 93%, 96%, and 98% percent of that of a maximally rugged random landscape, which, on average [mean±s.d.], has 2,308±133, 2,405±89 and 2,373±97 peaks for CRP, Fis and IHF, respectively. This analysis underlines that our landscapes are indeed highly rugged.

Because we use only quality-filtered genotype data, our landscapes lack regulatory information for about 40% of the 48 genotypes. While we cannot exclude that this undersampling of genotypes has led to systematic biases in our determination of landscape ruggedness, we note that the sampling of landscape genotypes by our experiments was not strongly biased with respect to regulation strength. Specifically, when we analyzed the relative connectivity of genotypes in our landscapes – the fraction of each genotype’s 24 possible neighbors for which our experiments yielded regulatory data – we found that it is only weakly correlated with regulation strength (R=-0.1, −0.1, 0.01 for the CRP, Fis, and IHF landscapes, Figures S13-S15). Similarly, the relative connectivity of peak genotypes is only weakly correlated with their regulation strength (R=-0.05, −0.04, 0.06 for the CRP, Fis, and IHF landscapes). Furthermore, the prevalence of sign epistasis (Supplementary Table S5) supports the notion that our landscapes are indeed rugged (see Supplementary Methods 7.5 for further details on epistasis and its evolutionary consequences).

Landscape peaks are widely scattered in genotype space

In a rugged adaptive landscape, reaching high fitness peaks can be challenging, because such peaks are separated from other genotypes by valleys of low fitness that cannot be traversed by natural selection alone38. If natural selection favors strong gene regulation, the ruggedness of our regulatory landscapes may thus present a challenge for adaptive evolution. To better understand the potential magnitude of this challenge, we next analyzed our landscapes’ topography in greater detail. We began by studying the distribution of peaks in genotype space. If a landscape’s highest peaks are widely scattered through genotype space, then they may be accessible via fitness-increasing evolutionary paths from diverse non-peak genotypes. This may facilitate adaptive evolution compared to a landscape where peaks are clustered in a small region of genotype space.

For each of our three landscapes, we determined the genetic distance between peaks, i.e., the minimum number of mutations needed to transition from one peak to another, regardless of their effect on regulation strength. We compared the distribution of these distances to the distribution of genetic distances for an equal number of randomly selected non-peak variants. In all three landscapes, the distances between peaks are almost indistinguishable from those of random genotypes, differing on average by fewer than 0.1 mutational steps. In other words, peaks are about as widely dispersed in each landscape as random genotypes (Supplementary Figure S16). This is the case for both low and high peaks (Supplementary Figures S17-S18). A principal component analysis further underscores this dispersion (Supplementary Methods 7.7, Supplementary Figures S19-S21).

Accessible paths to a peak are often not the shortest possible paths

Next, we focused on mutational paths to peaks that are evolutionarily accessible, i.e., a series of mutational steps where each step increases regulation strength. Specifically, we enumerated all accessible paths that exist from each non-peak genotype to each peak genotype for all three landscapes. We found that these paths are generally longer than the shortest genetic distance between a non-peak and the peak genotypes. Specifically, the mean length of accessible paths exceeded the shortest distance by 1.5, 1.8 and 1.8 steps for the CRP, Fis and IHF landscapes (Supplementary Figure S22). In other words, accessible mutational paths are often not the most direct paths.

The existence of indirect paths implies that some mutational steps are evolutionarily prohibited because they decrease regulation strength. The reason is closely linked to epistasis – non-additive effects of two or more mutations on regulation strength. (see Supplementary Methods 7.5 for further details on epistasis). More specifically, such inaccessible steps are a result of sign epistasis. In this kind of epistasis, a double mutant of a TFBS regulates expression more strongly than the TFBS itself, even though one or both constituent single mutants regulate expression more weakly than the TFBS37,75–77. Indeed, epistasis is prevalent in all three landscapes (Table S5). Specifically, we observe sign epistasis in 62%, 66%, and 65% percent of interactions between single mutant pairs in the CRP, Fis, and IHF landscapes.

The existence of accessible paths alone does not tell us how easily a high peak can be found through Darwinian evolution. The reason is that only a very small fraction of all paths to that peak may be accessible, and an evolving population may not find any one of those paths. To quantify the fraction of accessible paths, we determined the total number of paths from each non-peak genotype to each high fitness peak, and computed the fraction of those paths that are accessible. Figure 3a shows how this fraction depends on the number of mutational steps in a path. It behaves similarly for all three landscapes (Figure 3a). The majority of two-step paths (a fraction greater than 80%) are accessible in all three landscapes (Figure 3a), but the fraction of accessible paths dwindles rapidly with an increasing number of mutational steps. It reaches a minimum below 1% for all three landscapes at eight mutational steps (Figure 3a).

Peaks and their basins of attractions.

a. The fraction of accessible paths declines with increasing path length. The figure illustrates how the fraction of accessible paths to high peaks (vertical axis, logarithmic scale) decreases with the length of the shortest accessible path (horizontal axis). Accessible paths are defined as paths where each step increases regulation, starting from a specified initial genotype. Each colored line corresponds to data from a different TF landscape: CRP (green), Fis (orange), and IHF (blue). Circles indicate the mean fraction of accessible paths for a given path length. b. Higher peaks have larger basins of attraction. The split-violin plots display the distribution of basin sizes (vertical axis) for high (red) and low (blue) peaks in the three TF landscapes CRP, Fis, and IHF (horizontal axis). High peaks have significantly larger basins of attraction (Welch Two Sample t-tests; CRP: t-value = 9.0898, df = 63.509, and p-value = 4.22 × 10-13, with mean basin sizes of xhigh = 9528.426 and xlow = 5655.035. Fis: t-value = 7.617, df = 172.35, and p-value = 1.645 × 10-12, with mean basin sizes of xhigh = 14206.70 and xlow = 10931.94. IHF: t-value = 6.1777, df = 201.99, and p-value = 3.521 × 10-9, with mean basin sizes of xhigh = 10965.872 and xlow = 8664.953.) The violin plots show the distribution of basin sizes for each of the two kinds of peaks. Their width represents the frequency of a given basin size. The vertical length of the box in each plot covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR. c. The regulation strength of peak genotypes (horizontal axis) is plotted against the size of their basins of attraction (vertical axis, shown as a percentage of all (non-peak genotypes) for peaks in all three landscapes (color legend). Color-coded values of R at the top of the graph indicate Pearson correlation coefficients for each landscape (color legend). Curved lines are based on a linear regression analysis (in linear space), with grey shading indicating 95% confidence intervals (CRP: R2=0.28, N = 2,154; Fis: R2=0.24, N = 2,312; IHF: R2=0.23, N = 2,434). d. Basins of attractions share many TFBS genotypes.

High peaks have large basins of attraction that share many TFBS genotypes

In the following analysis, we computed the basin of attraction of each peak, defined as the set of non-peak genotypes from which accessible paths to the peak exist70,78,79. In other words, a peak’s basin of attraction comprises all genotypes from which adaptive evolution can access the peak. The peaks in our landscapes have widely different basin sizes. They include peaks accessible from a mere fraction of genotypes to those accessible by a majority of genotypes (Figure 3b). Notably, high peaks generally had larger basins of attraction than low peaks (Figure 3c, Supplementary Figure S23). In addition, we found that many genotypes are part of multiple basins of attraction. Adaptive evolution may reach different high peaks starting from any such genotype. For all pairs of high peaks, we computed the pairwise overlap (intersection) of the basins of attraction, i.e., the fraction of genotypes that are part of both basins of attraction. (Figure 3d, Supplementary Figures S24-S25). The number of genotypes in this intersection varies widely, but the basins of attraction of high peaks generally share a substantial fraction of genotypes (mean shared genotypes: 48%, 42%, and 36% among all pairs of high peaks for the CRP, Fis, and IHF landscapes, Figure 3d)

The violin plots with embedded boxplots illustrate the fraction of genotypes shared between basins of attractions (vertical axis) for all pairs of high peaks in the CRP (green), Fis (orange), and IHF (purple) TF landscapes (horizontal axis). Specifically, we computed basin overlap as one minus the Jaccard coefficient (Supplementary Methods 7.5). A value of one (zero) indicates that the basin of attraction of two peaks share all (no) genotypes (CRP: 61 high peaks, N=1,830 comparisons, Fis: 172 high peaks, N=14,706 comparisons, IHF: 199 high peaks, N=19,701 comparisons). Their width represents the frequency of a given basin overlap. The vertical length of the box in each box covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR.

Genetic drift facilitates and clonal interference reduces the attainment of high peaks

Taken together, our analyses thus far indicate that high peaks in the CRP, Fis, and IHF landscapes are more accessible than low peaks (Figure 3b), and their basins of attraction also share more genotypes (Figure 3d). However, the accessible evolutionary paths to high peaks are often indirect. In addition, the fraction of accessible paths to any one high peak dwindles rapidly with the distance from that peak (Figure 3a).

To understand how these topographical features jointly affect adaptive evolution, we simulated the evolutionary dynamics on our three landscapes under the assumption that natural selection would favor strong regulation and that fitness is proportional to regulation strength.

Because high peaks constitute only a small fraction of all peaks (2.8%, 0.4%, and 0.5% in the CRP, Fis, and IHF landscapes), we hypothesized that most evolving populations would reach only low peaks. To test this hypothesis, we took advantage of the fact that E.coli has a low genomic mutation rate of 2.2 × 10−10 per base pair per generation 43, and that we study evolution only within an 8-nucleotide segment of a TFBS. In addition, E. coli has a large effective population size (>108 individuals43), which means that genetic drift is weak and even minor differences in fitness are visible to natural selection 43,80. These conditions imply that a population on our landscape would evolve in the well-studied strong-selection weak-mutation regime (SSWM)39,81–83. In this regime, a population is monomorphic most of the time, until a beneficial mutation arises, which usually sweeps rapidly to fixation. In other words, adaptive evolution can be modeled as an adaptive walk, in which each mutational step is taken with a fixation probability that has been derived by Kimura 80. We thus call the resulting adaptive walks Kimura walks (Supplementary Methods 8). We simulated 103 such adaptive walks starting from each of 15,000 randomly and uniformly selected starting (non-peak) TFBS genotypes from each landscape. Each random walk comprised up to 25 mutational fixation steps, unless it reached a fitness peak earlier. Each adaptive walk also accounted for known E. coli mutational biases 84–86 (Supplementary Methods 8).

As we hypothesized, most adaptive walks indeed reach only low peaks (90%, 85%, and 87% percent for the CRP, Fis, and IHF landscapes.). However, the percentage of walks that terminate at high peaks is several-fold greater than the proportion of high peaks. Specifically, 10 percent of walks terminate at high peaks in the CRP landscape, which is 3.6-fold higher than the percentage (2.8%) of high peaks in this landscape. In the Fis and IHF landscape, adaptive walks are 2 and 1.6-fold more likely to terminate at high peaks than expected from the proportion of these peaks among all peaks (7.4% and 8.1% percent) (Figure 4a). In other words, adaptive walks reach high fitness peaks more often than one might expect.

Peak accessibility, contingency and biases in adaptive walks.

a. More than 10% percent of adaptive walks lead to high peaks. The graph shows the cumulative distribution function (CDF, vertical axis) of regulation strength (horizontal axis) attained at the end of 15 million adaptive walks (15,000 starting genotypes × 1,000 adaptive walks each) for each TF landscape (color legend, CRP=: green; Fis: orange; IHF: purple). Dashed lines intersect the CDF at points equivalent to the wild-type regulation strength for each TF (color legend). They help to infer the proportion yTF of walks that terminate at high peaks (S>SWT), which is indicated by the numerical values at the graph’s top for each of the landscapes (color legend). b. Paths to high peaks are not much longer than genetic distances from the starting genotype. Colored boxplots summarize, for each of the three landscapes (horizontal axis), the distribution of shortest genetic distances (blue) and actual path lengths (green) for adaptive walks terminating at high peaks. Path lengths are typically less than one mutational step longer than genetic distances. Each box covers the range between the first and third quartiles (IQR). The horizontal line within each box represents the median value, and whiskers span 1.5 times the IQR. Numbers atop each plot indicate means ± one standard deviation. c. Different numbers of high peaks are reached from different starting variants. Violin plots integrated with boxplots show the distribution in the number of high peaks reached (vertical axis) from each starting variant that attained any high peak during 10,000 adaptive walks. The number of these variants is shown as an integer on top of the panel, and its percentage among all 15,000 starting variants is shown in parentheses (color legend). Underneath it, the panel shows the mean ± 1 s.d. of the number of attained high peaks. a. d. Some high peaks are reached more often than others. We randomly and uniformly sampled 10 starting genotypes from the CRP landscape, started 103 adaptive walks from each, and recorded the number and frequency of distinct high peaks attained in these random walks. Results for each starting genotype are symbolized by a vertical bar. The number of stacks within each bar (delineated by horizontal lines, also indicated by an integer above each bar) indicates the number of high peaks reached by the 103 adaptive walks. Starting variants are ordered in ascending order based on this number of attained peaks. Stack height indicates the fraction of walks that reached the same peak, and is indicated in red, orange, and yellow for the three most frequently attained peaks. See Supplementary Figure S32 for the Fis and IHF landscapes.

The adaptive walks that reached a high peak were short (Figure 4b). On average, they were also no more than half a mutational step longer than the shortest genetic distance between each starting genotype and peak (Figure 4b, CRP landscape: path length 3.2 ± 1.6 [mean±s.d.] vs. genetic distance 2.8 ± 1.2; Fis landscape: path length 3.1 ± 1.5 vs. genetic distance 2.7 ± 1.2; IHF landscape: path length 3.1 ± 1.5 vs. genetic distance 2.7 ± 1.2).

These short paths can be explained by our previous observation that peaks are widely distributed across the genotype spaces of the three landscapes. This distribution makes it easier for populations starting from any (non-peak) genotype to find a local peak through few mutations. Additionally, path lengths realized during adaptive walks tend to be shorter than the average lengths of accessible paths. (Figure 4b) The reason is that the average length of accessible paths takes all accessible paths into consideration, whereas adaptive walks tend to stop at the first peak they encounter, which results in shorter realized path lengths.

Because genetic drift can cause evolving populations to escape a low fitness peak and attain a higher fitness peak, we also asked how small population sizes affect the likelihood that a population attains a high fitness peak (Supplementary Material 8, Supplementary Figures S26-28). When simulating adaptive evolution in populations of N=15,000 individuals, we found that the likelihood that a population reaches a high peak increased by at least 10% percent (to 18%, 20%, and 21% for the CRP, Fis, and IHF landscapes, Supplementary Figures S26-S28).

We also studied how deviations from the strong selection weak mutation regime would affect evolutionary dynamics. In large populations or at high mutation rates, populations tend to be polymorphic and are subject to clonal interference, where the most advantageous of several co-occurring mutations typically dominates and becomes fixed 87–89. To approximate this dynamic, we conducted simulations using “greedy” adaptive walks (Supplementary Methods 8) 38,90. In such an adaptive walk, it is always a genotype’s mutational neighbor with the highest increase in regulation strength that becomes fixed 82,90–92. In other words, a greedy walk starting from a given genotype is deterministic. We, therefore, simulated only one walk for each of the 15,000 randomly chosen (non-peak) starting genotypes. We found that the fraction of greedy walks reaching high regulation peaks is slightly lower than in the SSWM regime, with 1%, 2%, and 5% fewer walks achieving such peaks in the CRP, Fis, and IHF landscapes, respectively (Supplementary Figure S29). This outcome is expected, because greedy walks prioritize immediate fitness gains at the expense of the ability to discover higher fitness peaks90.

The attainment of peaks is highly contingent on chance events

Because different basins of attraction share many TFBS genotypes (Figure 3d, Supplementary Figures S24-S25), it is possible that adaptive evolution starting from any one genotype can reach different peaks, depending on which mutational paths it takes. In other words, the structure of the landscapes we study may lead to evolutionary contingency 93,94 – the dependence of an outcome of evolution on unpredictable chance events26,93,95. Indeed, non-peak genotypes can access on average around 30% of the total number of high peaks in each landscape (Supplementary Figure S25). To assess the prevalence of evolutionary contingency, we determined how many different peaks are attained by 103 adaptive walks starting from the same randomly chosen genotype. We performed this analysis on 15,000 starting genotypes, focusing on the subset of those starting genotypes from which at least one high peak is reached during the 103 walks. Specifically, 28.6%, 23%, and 24.5% of starting genotypes for the CRP, Fis, and IHF landscapes, reached at least one high peak. Notably, these percentages represent the fraction of starting genotypes that can access at least one high peak, not the total fraction of adaptive walks that lead to a high peak, which is lower (as shown in Figure 4a). For instance, while adaptive walks starting from 28.6% of genotypes reached at least one high peak in the CRP landscape only 10% of all adaptive walks in this landscape end at a high peak (Figure 4a). Importantly, for any one starting genotype where random walks reached at least one high peak, they typically reached more than one high peak (Figure 4c, median±IQR, 4±4, 4±8, 6±6 for the CRP, Fis, and IHF landscapes, respectively). The number of high peaks attained from any one starting genotype was as large as 21, 30, and 33 for the three respective landscapes. Thus, evolutionary contingency is pervasive in all three landscapes. In addition, not only the end point of evolution but also the evolutionary trajectories leading to this end point show contingency. That is, adaptive walks starting from a specific genotype and ending at a specific high peak often take multiple paths to this peak (Supplementary Figure S30).

Biases towards specific evolutionary outcomes have been documented in both empirical and computational studies 26,27,70,93,96–98. They also exist in our landscapes (Figure 4d, Supplementary Figure S31). Figure 4d illustrates both contingency and bias for 103 adaptive walks starting from each of 10 different genotypes in the CRP landscape (thus a total of 104 adaptive walks). Depending on the starting variant, the adaptive walks reached between 1 and 16 high peaks. Whenever different peaks are reached, they are reached by markedly different proportions of adaptive walks. The same holds in the Fis and IHF landscapes (Supplementary Figure S32). Additionally, while multiple routes to each peak can be traversed (Supplementary Figure S30), we observed in all three landscapes biases towards specific paths that are taken more frequently than others (Supplementary Figure S33).

Discussion

We evaluated the ability of three E. coli global regulators to control gene expression through each of more than 30,000 TFBSs for each regulator. To achieve this, we utilized a synthetic plasmid-based system that facilitates high-throughput fluorescence measurements. This system allowed us to quantify the regulatory strength of individual TFBSs by measuring gene expression through GFP fluorescence 51. Additionally, the system insulates the control of transcription from any direct effects library sequences might have on transcription, such as the transcriptional and translational impacts of 5’ untranslated regions on gene expression 99,100. Recent studies have investigated large empirical datasets of cis-regulatory genotypes, examining the impact of sequence variation on gene regulation in both eukaryotes and prokaryotes6,39,49,59,101–104. However, few studies have focused on these interactions from an evolutionary perspective, as we do here 6,39,103.

We showed that the regulatory landscapes of all three TFs are highly rugged and have multiple peaks. The ruggedness of all three landscapes is also supported by the prevalence of epistasis in them (Supplementary Table S5). A particularly important form of epistasis is sign epistasis37,75,76, because it can lead to multiple adaptive peaks 37,75,76 (see Supplementary Methods 7.5) Our landscapes contain up to 65% of mutation pairs with sign epistasis, a value that is especially high compared to the almost exclusively additive interactions of mutations in eukaryotic TFs6,105. However, the TFs we study are not exceptions among other prokaryotic TFs. Recent studies have shown that epistatic interactions are common in binding sites of local prokaryotic regulators, such as AraC (sign epistasis in more than 50% of 20 mutants106) and Cl (sign epistasis in 85% of 113 mutants107). We attribute this greater incidence of epistasis to the nature of prokaryotic TFBSs. Specifically, prokaryotic TFBSs are at approximately 20bps, twice as long as eukaryotic ones57,108 and exhibit symmetries reflecting the dimeric state of their cognate TFs109–111. These factors may increase the likelihood of intramolecular epistasis.

Despite high ruggedness and epistasis, we found that a modest proportion of evolving populations can access the highest peaks of each landscape. Specifically, in the large populations that are typical of E. coli, at least 10% percent of evolving populations reach one of the highest peaks. The clonal interference that may occur in even larger populations or in populations with high mutation rates only modestly reduces this percentage to 5% (Supplementary Figure S29) Conversely, in small populations where genetic drift can help a population escape from a low peak, this percentage increases to 18%. Numbers like these render the de novo evolution of strong transcription factor binding sites plausible for our three global regulators. Once such binding site have originated in one population, they can also spread to others through horizontal gene transfer112,113.

It is well known that epistasis can influence the trajectories of evolving populations on adaptive landscapes37,76,114. The regulation strength of a TFBS is partially determined by the combined interactions of the nucleotides within the TFBS115–118, which in turn affects TF-TFBS binding affinities. Different combinations of mutations can result in similar regulation strengths and create multiple evolutionary pathways to achieve optimal or near-optimal regulation6,119. This can lead to overlapping basins of attraction among different peak TFBSs conveying strong regulation, such that even evolution starting from the same genotypes can take different paths and reach different peaks with similar regulation strengths (Supplementary Figure S30). Indeed, we observed such overlapping basins in our landscapes (Figure 3d and Supplementary Figure S24). Moreover, epistasis can create plateaus of regulation strength where various combinations of nucleotides yield TFBSs with intermediate regulation strengths. These plateaus serve as common evolutionary intermediates that multiple starting sequences can traverse on their way to different peaks, and thus contribute further to the shared basins of attraction we observe (Supplementary Figure S24).

In a landscape with substantial epistasis, the sequence of mutations that occurs in an evolving population can render adaptive evolution highly contingent on this sequence93,94. For example, some beneficial mutations within a TFBS might only be accessible after other specific mutations have occurred. Different starting sequences or early mutations can also change the spectrum of accessible mutations, leading to different peaks with similarly strong regulation 93,94. Indeed, we observed in all three landscapes that different evolving populations starting from the same genotypes in the landscape attain different peaks (Figures 3d, 4d and Supplementary Figure S32) 93,95,120,121. Such contingency renders evolution less predictable 26,93,122.

Despite the prevalence of contingency in all three landscapes, evolving populations starting from the same genotype more often attain some peaks than others (Supplementary Figure S32). Moreover, for each attained peak with multiple possible evolutionary routes, we also observed that some paths are more frequently transversed than their alternatives (Supplementary Figure S33). That is, evolution is biased towards traversing some paths and attaining some peaks more often than others123,124. These observations emphasize the complex interplay between chance, contingency, and evolutionary biases in shaping the outcomes of adaptive evolution.26,93,95,97.

The three TFs we studied here belong to different protein families, yet they have regulatory landscapes with similar topography. All three landscapes highly rugged, highly epistatic, and harbor multiple peaks that are widely scattered, and this holds for peaks of low, intermediate, and high regulation strength. This might help to explain how TFBSs with intermediate and high regulation strength can be easily attained during evolution. In addition, peak accessibility in all three landscapes increases with peak height (Supplementary Figure S31). On the one hand, these commonalities may be caused by common biological or biochemical properties of the three TFs. For example, CRP has been suggested to possess nucleoid-associated protein properties similar to Fis and IHF, due to its ability to bend and loop DNA125. On the other hand, the commonalities may reflect general characteristics of global gene regulation. One of them is that global TFs often bind unspecifically to multiple TFBSs11,13. Also, they may bind DNA with a broad range of different affinities centered around intermediate affinity, rather than bind few sites but very strongly 13. This possibility is supported by comprehensive analyses of in vitro eukaryotic TF binding affinities for thousands of TFBS variants, showing essentially continuous binding affinity distributions for hundreds of sites6.

Despite broad similarities among the three landscapes we study, we also found some differences. Most notable is the narrower distribution of regulation strengths for the binding sites of CRP compared to those of Fis and IHF (Figure 2a). This is not unexpected, given that CRP binding sites are known for being quasi-symmetric and less degenerate than those of Fis and IHF28,31. Previous studies have also shown that Fis and IHF binding sites are less conserved in their nucleotide structure and more biased toward AT-rich content22,126,127. One possible explanation for this observation is the nature of the binding dynamics of these regulators. CRP, mostly restricted to gene regulatory interactions, may have evolved to finely tune its regulatory output, resulting in a narrower distribution of regulation strengths. In contrast, Fis and IHF, which are involved in functions beyond gene regulation, such as DNA replication and genomic structural maintenance128,129, might encompass a more diverse range of regulation strengths.

The landscapes of our global regulators also differ in two respects from that of the previously characterized local regulator TetR 27. First, the peaks of the TetR landscape are more clustered in genotype space due to its lower diversity of strongly binding TFBSs 27. Second, the distribution of regulation strengths for its TFBSs is narrower and shows more strongly binding TFBSs 27. A possible explanation may lie in the fact that TetR regulates only two transcription units130.

Some of the properties of our landscapes are similar to what would be expected from theoretical models, such as uncorrelated random landscapes72,73,131,132. First, both kinds of landscapes have many peaks. Specifically, the number of peaks of the CRP, Fis, and IHF landscapes is 93%, 96%, and 98% of that expected in a random landscape of the same size (Supplementary methods 7.6). Second, the peaks of our landscapes are widely scattered in genotype space. This scattering may reflect a biological property, namely that global regulators can strongly bind multiple and highly diverse sites11–13. In addition, high peaks in our empirical landscapes also have large basins of attractions, an observation that is consistent with predictions from simple theoretical models 132.

One limitation of our work stems from our rigorous quality filtering of sort-seq data. As a result, we lack regulatory data for some 40% of the TFBSs in each landscape. This limited diversity of reliable data is a common feature in mutational library studies. It has several technical causes, such as biases in library synthesis 133, PCR amplification 134, cloning, and loss of sequence diversity after cell sorting 102. In future work, this limitation could be overcome by a combination of strategies, such as subsampling complete genotype spaces or combining different molecular methods to overcome biases and diversity loss during PCR amplification135,136, sorting48,137,138, and high-throughput sequencing103.

A second limitation comes from our use of the sort-seq method, which is best-suited for our work, because it allows the high-throughput measurement and sorting of millions of individual cells in a standardized and straightforward manner. However, the method’s accuracy depends on the binning procedure. Other studies have used between two and several dozen bins, with or without unbiased sampling 48,50,59,101,102,137,139. Recommendations emerging from this work include sorting of cells into at least four logarithmically (log2) equally-spaced bins, each covering approximately 12.5-15% of the fluorescence distribution, 59,102,140. We followed these recommendations, using 13 bins. In addition, we computed a weighted average to calculate expression values from sequences appearing in multiple bins, a straightforward method validated by robust studies in transcriptional regulation analysis 39,49. In addition, we validated individual regulation strengths with an independent method to demonstrate its reliability.

A third limitation of our study is that we only examined variation at eight positions for each TF. We selected these positions based on their importance for DNA binding and regulation by our TFs, as evidenced by their high information content. We cannot exclude the possibility that selecting a different set of positions could yield different landscape topographies. However, we speculate that less information-rich position would not reduce but rather increase the potential for the de novo evolution of strong TFBSs. For example, they may facilitate landscape navigability through extradimensional bypasses78,141 or provide small-incremental changes in regulation strengths. These small changes help populations reach peaks via diminishing returns effects, meaning that as a population gets closer to a peak, each subsequent mutation contributes progressively smaller improvements to regulation strength 142,143. Investigating a wider array of positions and larger regulatory landscapes remains an important task for future work.

Fourth, we use a simplified empirical system for the study of gene regulation—a promoter followed by a single TFBS. Although this regulatory architecture exists for some genes144, global regulators often form part of more complex, combinatorial architectures4,144,145 that are influenced by environmental factors2,146,147, concentrations of active TFs148–150, chromosome structure151–155, methylation states156, and more. Studies on more complex promoter architectures may reveal regulatory landscapes with different topographies.

Lastly, we also analyzed the adaptive evolution of TFBSs in a simplified manner, performing data-driven evolutionary simulations rather than experimental evolution. Although many studies assume that regulation strength and fitness are correlated, this is not always the case. Low binding affinities can be adaptive during development 157, and many genes exhibit a nonlinear fitness-expression function 158 with a plateau of maximal fitness across a wide range of expression levels 159–161. For instance, most mutations and polymorphisms in the promoter of the yeast gene TDH3 do not significantly affect fitness in a glucose-rich medium160. For these reasons we focused our observations and interpretations on regulatory phenotypes (regulation strength). Our simplified approach allowed us to model population evolution in a large genotype space and avoid monitoring thousands of evolving regulatory sequences simultaneously in vivo162,163. However, it cannot replace experimental evolution of TFBSs, which also remains an important challenge for future work.

Transcription factor binding sites are among the simplest units of biological organization. Our work provides the first large-scale analysis of the regulatory landscapes formed by such sites for global transcriptional regulators. It shows that strong binding sites can readily evolve de novo, even though prokaryotic transcription factor binding sites are much larger than their eukaryotic counterparts and have more rugged regulatory landscapes. In addition, the evolution of these simple sequences also displays phenomena that have been characterized in much more complex systems, such as evolutionary contingency and evolutionary biases.

Materials and methods

The Supplementary Materials contain extended details of experimental procedures and data analysis.

Strains and plasmids

Bacterial strains and plasmids used in this work are listed in Table 1. We obtained electrocompetent E. coli cells of strain SIG10-MAX® from Sigma Aldrich (CMC0004). We used this strain for molecular cloning and library generation due to its high transformation efficiency. The genotype of this strain (Supplementary Table S4) is similar to DH5α (Sigma Aldrich commercial information, see Supplementary Table S4). The strain is resistant to the antibiotic streptomycin.

We amplified plasmid libraries in SIG10-MAX®, extracted their DNA, and transformed them into mutants derived from E. coli K-12 strain BW25113 that harbor chromosomal deletions of the crp, fis, or ihfa gene. We obtained these mutant strains from the KEIO collection 164 and used them for sort-seq experiments. The design, genetic parts, and assembly of the plasmid vectors we used in this study are available in the Supplementary material, as are all primers, TFBS sequences/libraries, strains, and plasmids.

Sort-Seq procedure

To explore the regulatory effects of each transcription factor on binding sites in the corresponding library, we constructed three plasmids, each of which enables the inducible expression of one of our three TFs. These are plasmids pCAW-Sort-Seq-V2-CRP, pCAW-Sort-Seq-V2-Fis, and pCAW-Sort-Seq-V2-IHF (Supplementary Methods 3 and Supplementary Table S1). We cloned the TFBS libraries (Supplementary Table S3) into their respective plasmids and then transformed them into mutant strains lacking the corresponding TFs (Δcrp, Δfis, and Δihf, as listed in Supplementary Table S2). We induced TF expression using anhydrotetracycline (Atc) and, after overnight growth, performed cell sorting for cell populations. During sorting, we distributed cells into 13 equally-spaced logarithmic bins based on their fluorescence levels. We replicated each sort-seq experiment three times from three separate library transformations for each TF. To mitigate the impact of extrinsic noise (gene expression variation among cells 165), we adhered to standard protocols by normalizing our GFP fluorescence measurements against mScarlet-I fluorescence values obtained from flow-cytometry assays 7,48,166,167. We subsequently recovered the sorted cells from each bin in 50mL Falcon tubes containing 10 mL of LB medium supplemented with chloramphenicol and incubated them overnight at 37°C with shaking at 220 rpm. After this growth period, we re-sorted cell cultures from each bin to eliminate potential contaminants and ensure that the cell populations had preserved their fluorescence distributions. Following re-sorting, we extracted plasmids from the cell population of each bin. We amplified and barcoded the TFBS region from each population through a polymerase chain reaction (PCR). Lastly, we sequenced barcoded amplicons containing TFBS sequences, and used the sequencing results to calculate the regulation strengths of each TF to its TFBSs in the corresponding library. More details are provided in the Supplementary Material.

Regulation strengths

Due to gene expression and measurement noise, individual TFBS variants in a sort-seq experiment usually appear in more than a single bin, and their read count (frequency) varies among bins48,49,101,138. Following established practice27,49,103, we used a weighted average of these frequencies for each variant to represent the mean expression level driven by the variant. To facilitate the interpretation of this quantity, we converted this expression level into a regulation strength relative to the highest observed regulation strength for a given TF, to which we assigned a value of one. From each library, we selected a single naturally occurring binding site for each TF that was previously characterized in the literature as a strong binder. We called this TFBS the WT sequence and used it as a baseline to separate weakly (higher GFP expression) from strongly (lower GFP expression) binding TFBSs.

Validating regulation strengths with plate reader measurements

To further validate our regulation strength data, we chose 10 DNA binding sites from each bin (10 variants ×13 bins = 130 variants in total per TF library, Supplementary File 2), covering a wide range of measured regulation strengths. We cloned these sequences into the appropriate vector pCAW-Sort-Seq-V2-TF (TF: CRP, Fis or IHF), and transformed them into the appropriate mutant strain. We picked individual colonies and grew them overnight (16 hours, 37°C, 220 rpm) in liquid LB supplemented with 50 μg/mL of chloramphenicol and anhydrotetracycline. We diluted the cultures to 1:10 (v/v) in cold Dulbecco’s PBS (Sigma-Aldrich #D8537) to a final volume of 1 mL. We transferred 200 μl of the diluted cultures into individual wells in 96-well plates and measured GFP fluorescence (emission: 485nm/excitation: 510nm, bandpass: 20nm, gain: 50), as well as the optical density at 600nm (OD600) as an indicator of cell density. We then normalized fluorescence by the measured OD600 value to account for differences in cell density among cultures and compared the obtained ratios to the previously inferred regulation strengths for the 130 selected variants. We performed all such measurements in biological and technical triplicates (three colonies per sample, and three wells per colony, respectively).

Code Availability and Data Analysis

All code used for processing data and plotting, as well as the final processed data, plasmid sequences, and primer sequences are available in our GitHub repository.

eLife Assessment

This study maps the genotype-phenotype landscapes of three E. coli transcription factors and the topographical features of these landscapes. It shows that ruggedness and epistasis do not hinder the evolution of strong transcription factor binding sites. These convincing findings contribute valuable insights into fitness landscape theories and highlight the role of chance, contingency, and evolutionary biases in gene regulation. The authors then study the topographical features of these landscapes, especially the number and distribution of local maxima, as well as the statistical properties of evolutionary paths on these landscapes.

Reviewer #1 (Public review):

Summary:

For each of the three key transcription factor (TF) proteins in E. coli, the authors generate a large library of TF binding site (TFBS) sequences on plasmids, such that each TFBS is coupled to the expression of a fluorescence reporter. By sorting the fluorescence of individual cells and sequencing their plasmids to identify each cell's TFBS sequence (sort-seq), they are able to map the landscape of these TFBSs to the gene expression level they regulate. The authors then study the topographical features of these landscapes, especially the number and distribution of local maxima, as well as the statistical properties of evolutionary paths on these landscapes. They find the landscapes to be highly rugged, with about as many local peaks as a random landscape would have, and with those peaks distributed approximately randomly in sequence space. The authors find that there are a number of peaks that produce regulation stronger than that of the wild-type sequence for each TF and that it is not too unlikely to reach one of those "high peaks" from a random starting sequence. Nevertheless, the basins of attractions for different peaks have significant overlap, which means that chance plays a major role in determining which peak a population will evolve to.

Strengths:

(1) The experiments and analysis of this paper are very well-executed and, by and large, very thorough (with an important exception identified below). I appreciated the systematic nature of the project, both the large-scale experiments done on three TFs with replicates and the systematic analysis of the resulting landscapes. This not only makes the paper easy to follow but also inspires confidence in their results since there is so much data and so many different ways of analyzing it. It's a great recipe for other studies of genotype-phenotype landscapes to follow.

(2) Considering how technical the project was, I am really impressed at how easy to read I found the paper, and the authors deserve a lot of credit for making it so. They do a great job of building up the experiments and analyses step-by-step and explaining enough of the basics of the experimental design and the essence of each analysis in the main text without getting too complicated with details that can be left to the Methods or SI. Compared to other big data papers, this one was refreshingly not overwhelming.

Weaknesses:

(1) The main weakness of this paper, in my view, is that it felt disconnected from the larger body of work on fitness and genotype-phenotype landscapes, including previous data on TFBSs in E. coli, genotype-phenotype maps of TFBSs in other systems, protein sequence landscapes (e.g., from mutational scans or combinatorially-complete libraries), and fitness landscapes of genomic mutations (e.g., combinatorially-complete landscapes of antibiotic resistance alleles). I have no doubt the authors are experts in this literature, and they probably cite most of it already given the enormous number of references. But they don't systematically introduce and summarize what was already known from all that work, and how their present study builds on it, in the Abstract and Introduction, which left me wondering for most of the paper why this project was necessary. Eventually, the authors do address most of these points, but not until the end, in the Discussion. Readers who have no familiarity with this literature might read this paper thinking that it's the first paper ever to study topography and evolutionary paths on genotype-phenotype landscapes, which is not true.

There were two points that made this especially confusing for me. First, in order to choose which nucleotides in the binding sites to vary, the authors invoke existing data on the diversity of these sequences (position-weight matrices from RegulonDB). But since those PWMs can imply a genotype-phenotype map themselves, an obvious question I think the authors needed to have answered right away in the Introduction is why it is insufficient for their question. They only make a brief remark much later in the Results that the PWM data is just observed sequence diversity and doesn't directly reflect the regulation strength of every possible TFBS sequence. But that is too subtle in my opinion, and such a critical motivation for their study that it should be a major point in the Introduction.

The second point where the lack of motivation in the Introduction created confusion for me was that they report enormous levels of sign epistasis in their data, to the point where these landscapes look like random uncorrelated landscapes. That was really surprising to me since it contrasts with other empirical landscape data I'm familiar with. It was only in the Discussion that I found some significant explanation of this - namely that this could be a difference between prokaryotic TFBSs, as this paper studies, and the eukaryotic TFBSs that have been the focus of many (almost all?) previous work. If that is in fact the case - that almost all previous studies have focused on eukaryotic TFBSs or other kinds of landscapes, and this is the first to do a systematic test of prokaryotic TFBS, then that should be a clear point made in the Abstract and Introduction. (I find a comparable statement only in the very last paragraph of the Discussion.) If that's the case, then I would also find that point to be a much stronger, more specific conclusion of this paper to emphasize than the more general result of observing epistasis and contingency (as is currently emphasized in the Abstract), which has been discussed in tons of other papers. This raises all sorts of exciting questions for future studies - why do the landscapes of prokaryotic TFBSs differ so dramatically from almost all the other landscapes we've observed in biology? What does that mean for the evolutionary dynamics of these different systems?

(2) I am a bit concerned about the lack of uncertainties incorporated into the results. The authors acknowledge several key limitations of their approach, including the discreteness of the sort-seq bins in determining possible values of regulation strength, the existence of a large number of unsampled sequences in their genotype space, as well as measurement noise in the fluorescence readouts and sequencing. While the authors acknowledge the existence of these factors, I do not see much attempt to actually incorporate the effect of these uncertainties into their conclusions, which I suspect may be important. For example, given the bin size for the fluorescence in sort-seq, how confident are they that every sequence that appears to be a peak is actually a peak? Is it possible that many of the peak sequences have regulation strengths above all their neighbors but within the uncertainty of the fluorescence, making it possible that it's not really a peak? Perhaps such issues would average out and not change the statistical nature of their results, which are not about claiming that specific sequences are peaks, just how many peaks there are. Nevertheless, I think the lack of this robustness analysis makes the results less convincing than they otherwise would be.

Reviewer #2 (Public review):

The authors aim to investigate the ability of evolution to create strong transcription factor binding sites (TFBSs) de novo in E. coli. They focus on three global transcriptional regulators: CRP, Fis, and IHF, using a massively parallel reporter assay to evaluate the regulatory effects of over 30,000 TFBS variants. By analyzing the resulting genotype-phenotype landscapes, they explore the ruggedness, accessibility, and evolutionary dynamics of regulatory landscapes, providing insights into the evolutionary feasibility of strong gene regulation. Their experiments show that de novo adaptive evolution of new gene regulation is feasible. It is also subject to a blend of chance, historical contingency, and evolutionary biases that favor some peaks and evolutionary paths.

(1) Strengths of the methods and results:

The authors successfully employed a well-designed sort-seq assay combined with high-throughput sequencing to map regulatory landscapes. The experimental design ensures reliable measurement of regulation strengths. Their system accounts for gene expression noise and normalizes measurements using appropriate controls.

Comprehensive Landscape Mapping:

The study examines ~30,000 TFBS variants per transcription factor, providing statistically robust and thorough maps of the regulatory landscapes for CRP, Fis, and IHF. The landscapes are rigorously analyzed for ruggedness (e.g., number of peaks) and epistasis, revealing parallels with theoretical uncorrelated random landscapes.

Evolutionary Dynamics Simulations:

Through simulations of adaptive walks under varying population dynamics, the authors demonstrate that high peaks in regulatory landscapes are accessible despite ruggedness. They identify key evolutionary phenomena, such as contingency (multiple paths to peaks) and biases toward specific evolutionary outcomes.

Biological Relevance and Novelty:

The author's work is novel in focusing on global regulators, which differ from previously studied local regulators (e.g., TetR). They provide compelling evidence that rugged landscapes are navigable, facilitating de novo evolution of regulatory interactions. The comparison of landscapes for CRP, Fis, and IHF underscores shared topographical features, suggesting general principles of global transcriptional regulation in bacteria.

(2) Weaknesses of the methods and results:

Undersampling of Genotype Space:

While the quality filtering of the data ensures robustness, ~40% of the TFBS space remains uncharacterized. The authors acknowledge this limitation but could improve the analysis by employing subsampling or predictive modeling.

Simplified Regulatory Architecture:

The study considers a minimal system of a single TFBS upstream of a reporter gene. While this may have been necessary for clarity, this simplification may not reflect the combinatorial complexity of transcriptional regulation in vivo.

Lack of Experimental Validation of Simulations:

The adaptive walks are based on simulated dynamics rather than experimental evolution. Incorporating in vivo experimental evolution studies would strengthen the conclusions. Although this is a large request for the paper, that would not prevent publication.

Impact on the Field:

This study advances our understanding of adaptive landscapes in gene regulation and offers a critical step toward deciphering how global regulators evolve de novo binding sites. The findings provide foundational insights for synthetic biology, evolutionary genetics, and systems biology by highlighting the evolutionary accessibility of strong regulation in bacteria.

Utility of Methods and Data:

The sort-seq approach, combined with landscape analysis, provides a robust framework that can be extended to other transcription factors and systems. If made publicly available, the study's data and code would be valuable for researchers modeling transcriptional regulation or studying evolutionary dynamics.

Additional Context:

The study builds on a growing body of work exploring regulatory evolution. For instance, recent studies on local regulators like TetR and AraC have revealed high ruggedness and epistasis in TFBS landscapes. This study distinguishes itself by focusing on global regulators, which are more biologically complex and influential in bacterial gene networks. The observed evolutionary contingency aligns with findings in other biological systems, such as protein evolution and RNA folding landscapes, underscoring the generality of these evolutionary principles.

Conclusion:

The authors successfully mapped the genotype-phenotype landscapes for three global regulators and simulated evolutionary dynamics to assess the feasibility of strong TFBS evolution. They convincingly demonstrate that ruggedness and epistasis, while prominent, do not preclude the evolution of strong regulation. Their results support the notion that gene regulation evolves through a blend of chance, contingency, and evolutionary biases.

This paper makes a significant contribution to the understanding of regulatory evolution in bacteria. While minor limitations exist, the authors' methods are robust, and their findings are well-supported. The work will likely be of broad interest to researchers in molecular evolution, synthetic biology, and gene regulation.

Author Response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Weaknesses:

(1) The main weakness of this paper, in my view, is that it felt disconnected from the larger body of work on fitness and genotype-phenotype landscapes, including previous data on TFBSs in E. coli, genotype-phenotype maps of TFBSs in other systems, protein sequence landscapes (e.g., from mutational scans or combinatorially-complete libraries), and fitness landscapes of genomic mutations (e.g., combinatorially-complete landscapes of antibiotic resistance alleles). I have no doubt the authors are experts in this literature, and they probably cite most of it already given the enormous number of references. But they don't systematically introduce and summarize what was already known from all that work, and how their present study builds on it, in the Abstract and Introduction, which left me wondering for most of the paper why this project was necessary. Eventually, the authors do address most of these points, but not until the end, in the Discussion. Readers who have no familiarity with this literature might read this paper thinking that it's the first paper ever to study topography and evolutionary paths on genotype-phenotype landscapes, which is not true.

There were two points that made this especially confusing for me. First, in order to choose which nucleotides in the binding sites to vary, the authors invoke existing data on the diversity of these sequences (position-weight matrices from RegulonDB). But since those PWMs can imply a genotype-phenotype map themselves, an obvious question I think the authors needed to have answered right away in the Introduction is why it is insufficient for their question. They only make a brief remark much later in the Results that the PWM data is just observed sequence diversity and doesn't directly reflect the regulation strength of every possible TFBS sequence. But that is too subtle in my opinion, and such a critical motivation for their study that it should be a major point in the Introduction.

The second point where the lack of motivation in the Introduction created confusion for me was that they report enormous levels of sign epistasis in their data, to the point where these landscapes look like random uncorrelated landscapes. That was really surprising to me since it contrasts with other empirical landscape data I'm familiar with. It was only in the Discussion that I found some significant explanation of this - namely that this could be a difference between prokaryotic TFBSs, as this paper studies, and the eukaryotic TFBSs that have been the focus of many (almost all?) previous work. If that is in fact the case - that almost all previous studies have focused on eukaryotic TFBSs or other kinds of landscapes, and this is the first to do a systematic test of prokaryotic TFBS, then that should be a clear point made in the Abstract and Introduction. (I find a comparable statement only in the very last paragraph of the Discussion.) If that's the case, then I would also find that point to be a much stronger, more specific conclusion of this paper to emphasize than the more general result of observing epistasis and contingency (as is currently emphasized in the Abstract), which has been discussed in tons of other papers. This raises all sorts of exciting questions for future studies - why do the landscapes of prokaryotic TFBSs differ so dramatically from almost all the other landscapes we've observed in biology? What does that mean for the evolutionary dynamics of these different systems?

We thank the reviewer for this thoughtful and detailed critique. We agree that the original version of the manuscript did not sufficiently motivate the study early on, nor did it clearly position our work within the broader literature on genotype–phenotype (GP) and fitness landscapes. We also agree that two specific issues, the role of PWMs and the unexpectedly high levels of sign epistasis, were insufficiently explained early on, which could lead to confusion for readers not already familiar with this field.

Positioning within the broader landscape literature

In response, we have substantially revised the Abstract and Introduction to explicitly situate our work within existing empirical studies of GP and fitness landscapes, including TFBS landscapes in bacteria, eukaryotic TFBS genotype–phenotype maps, in vitro TF–DNA binding studies, deep mutational scans of proteins, and combinatorially complete fitness landscapes such as antibiotic resistance alleles (Abstract; Introduction, lines 64–85). We now make clear that our study builds directly on this extensive body of work, rather than introducing the landscape framework itself. For example, we write in the introduction:

“Over the last decade, genotype–phenotype (GP) maps and fitness landscapes have become central tools for understanding how molecular systems evolve under mutation and selection[22–25]. Such maps and landscapes have been experimentally studied for DNA[6,8,18,19,26,27], protein[28–32] and RNA[33–35] molecules, revealing key topographical properties that shape evolutionary outcomes, including epistasis[24,36]—the non-additive effects of multiple mutations on phenotype—landscape ruggedness, reflected in the number and distribution of fitness peaks, and constraints on adaptive evolution.”

At the same time, we clarify what remains rare in the literature: large-scale, in vivo genotype–phenotype landscapes for bacterial transcription factor binding sites that are sufficiently dense to support explicit evolutionary analyses. While numerous high-throughput studies have characterized bacterial regulatory elements, these datasets typically do not provide quantitative regulatory phenotypes across large genotype spaces, nor do they analyze evolutionary accessibility. To our knowledge, only one such in vivo TFBS landscape had previously been characterized at comparable resolution for a bacterial local regulator (TetR). Our work extends this approach to three global regulators, enabling systematic comparisons across prokaryotic systems (Abstract, Introduction, lines 64–85). For example, we write in the introduction:

“For transcription factor binding sites, most pertinent large-scale studies are based on in vitro binding assays, such as protein-binding microarrays (PBMs), and they focus predominantly on eukaryotic transcription factors[6]. While these studies have been instrumental in characterizing transcription factor binding preferences, they typically do not measure regulatory output in a native cellular context. In contrast, comprehensive in vivo data for bacterial TFBSs remain extremely rare. To our knowledge, only two high-resolutionin vivo landscapes have been previously mapped for bacterial regulators, those of the local regulators TetR[18] and LacI[27]. As a result, it remains unclear whether principles inferred from protein landscapes, eukaryotic TFBSs, or in vitro binding assays generalize to transcriptional regulation in bacteria, particularly for global regulators[11] that integrate multiple physiological signals.”

Why PWMs are insufficient for our question.

We agree with the reviewer that our original explanation of the role of PWMs was too cursory and should have been addressed explicitly in the Introduction. We have now revised the Introduction to clearly explain why PWMs derived from RegulonDB cannot substitute for empirical GP landscapes in our study (Introduction, lines 102–113).

In this passage we now explain that, first, PWMs are inferred from a limited number of naturally occurring binding sites—typically on the order of hundreds of sequences—whose diversity reflects evolutionary history and genomic context rather than systematic exploration of sequence space. As a result, PWMs sample only a small and biased subset of the possible TFBS variants, whereas our libraries probe tens of thousands of sequences in a controlled manner, providing substantially broader and more uniform coverage of genotype space (Introduction, lines 102–113).

Second, PWM scores are not direct measurements of regulatory strength. Instead, they represent probabilistic or heuristic scores that are primarily used for identifying candidate binding sites in genomes. Numerous studies have shown that PWM scores often correlate weakly with in vivo binding affinity or regulatory output, where DNA shape, cooperative interactions, and chromosomal context play important roles. As such, PWMs do not provide quantitative genotype–phenotype relationships for regulation strength (Introduction, lines 102–113).

Third, PWMs assume independent and additive contributions of individual nucleotide positions. This assumption excludes epistatic interactions by construction. Because epistasis is central to landscape ruggedness, peak structure, and evolutionary accessibility, PWM-based models are fundamentally unsuited to address the evolutionary questions we study here (Introduction, lines 102–113). We now explicitly state this limitation early in the manuscript, rather than only alluding to it later in the Results.

Sign epistasis and contrast with prior TFBS landscapes.

We also agree with the reviewer that the extensive sign epistasis we observe—approaching levels expected for uncorrelated random landscapes—is surprising in light of much of the existing empirical landscape literature. Importantly, as the reviewer notes, most previous TFBS landscape studies have focused on in vitro binding systems or on eukaryotic transcription factors, which tend to exhibit smoother and more additive landscapes.

To address this concern, we have revised the Abstract and Introduction to explicitly frame this contrast as a central result of the study (Abstract; Introduction, lines 151-153, Discussion, lines 652–668). For example, we write in the discussion:

“We showed that the regulatory landscapes of all three TFs are highly rugged and have multiple peaks. The ruggedness of all three landscapes is also supported by the prevalence of epistasis between pairs of TFBS mutations (Supplementary Table S5). A particularly important form of epistasis is sign epistasis[24,93,94], because it can lead to multiple adaptive peaks [24,93,94] (see Supplementary Methods 7.5). Our landscapes contain up to 65% of mutation pairs with sign epistasis, a value that is especially high compared to the almost exclusively additive interactions of mutations in eukaryotic TFs[6,125].”

We now emphasize that prokaryotic TFBS landscapes, particularly for global regulators, appear to be substantially more rugged and epistatic than most previously characterized TFBS landscapes, and that this difference likely reflects fundamental biological distinctions between regulatory systems.

Revised emphasis and conclusions.

Following the reviewer’s suggestion, we have adjusted the emphasis of the manuscript accordingly. Rather than highlighting epistasis and contingency as generic evolutionary phenomena, we now present the extreme ruggedness of prokaryotic TFBS landscapes as a system-specific finding with important implications for the evolution of gene regulation. We explicitly note that this raises new questions for future work—such as why prokaryotic regulatory landscapes differ so markedly from eukaryotic ones, and how these differences shape evolutionary dynamics—which we now highlight in the Introduction and Discussion (Abstract; Introduction, lines 151-153, Discussion, lines 652–668). For example, we write in the discussion:

“… A possible reason for this greater incidence of epistasis lies in the nature of prokaryotic TFBSs. Specifically, prokaryotic TFBSs are at approximately 20bps twice as long as eukaryotic TFBSs[80,128] and exhibit symmetries that reflect the dimeric state of their cognate TFs[129–131]. These factors may increase the likelihood of intramolecular epistasis. Our observations raise important questions for future work, such as why the landscapes of prokaryotic TFBSs differ so dramatically from those of eukaryotic ones. And what do these differences imply for the evolutionary dynamics of gene regulation?”

We believe that these revisions substantially improve the clarity, motivation, and positioning of the manuscript, and directly address the reviewer’s concerns by making both the necessity and the novelty of the study clear from the outset.

(2) I am a bit concerned about the lack of uncertainties incorporated into the results. The authors acknowledge several key limitations of their approach, including the discreteness of the sort-seq bins in determining possible values of regulation strength, the existence of a large number of unsampled sequences in their genotype space, as well as measurement noise in the fluorescence readouts and sequencing. While the authors acknowledge the existence of these factors, I do not see much attempt to actually incorporate the effect of these uncertainties into their conclusions, which I suspect may be important. For example, given the bin size for the fluorescence in sort-seq, how confident are they that every sequence that appears to be a peak is actually a peak? Is it possible that many of the peak sequences have regulation strengths above all their neighbors but within the uncertainty of the fluorescence, making it possible that it's not really a peak? Perhaps such issues would average out and not change the statistical nature of their results, which are not about claiming that specific sequences are peaks, just how many peaks there are. Nevertheless, I think the lack of this robustness analysis makes the results less convincing than they otherwise would be.

We thank the reviewer for raising this important concern. We fully agree that uncertainties arising from experimental resolution, measurement noise in fluorescence and sequencing, and incomplete sampling of genotype space should be incorporated explicitly into the analysis. While these limitations were acknowledged qualitatively in the original manuscript, we recognize that a direct, quantitative assessment of their impact on our conclusions is essential to strengthen the robustness of the study.

We first clarify that regulation strength is not discretized in our analysis. For each TFBS, regulation strength is calculated as a continuous weighted average of fluorescence across all sorting bins, based on the sequencing read-count distribution of each sequence across bins. We clarified this information in the main text (Results, lines 201-203). Nevertheless, finite binning resolution and experimental noise introduce uncertainty in these estimates, which could in principle affect the identification of local peaks.

Importantly, our study does not aim to assert that specific TFBS sequences are definitively peaks. Rather, our focus is on landscape-level statistical and topological properties—such as ruggedness, the abundance and distribution of peaks, and the evolutionary accessibility of strong regulation. We therefore centered our new analyses on testing whether these conclusions are robust to experimentally plausible sources of uncertainty, rather than on the identity of individual peaks.

To address the reviewer’s concern, we performed two complementary analyses. The first evaluates whether the observed ruggedness of the landscapes could arise as an artifact of incomplete sampling. It addressed the effects of missing genotypes and the possibility of spurious peak identification due to unsampled neighbors. Sparse sampling can introduce opposing biases: true peaks may be missed, while other genotypes may be falsely classified as peaks because fitter neighbors are absent. As shown for uncorrelated random (House-of-Cards) landscapes (Kauffman & Levin, 1987), these effects can partially cancel.

In this analysis, we constructed a null model by randomly permuting regulation strengths across the mapped genotype network while preserving its topology. The number of peaks in these randomized landscapes is only modestly higher than in the empirical data, indicating that the measured landscapes are close to the maximal ruggedness compatible with the sampled network (Results, lines 308–320).

In addition, we quantified potential sampling bias by analyzing genotype connectivity. Here we defined the relative connectivity of a genotype as the fraction of possible single-mutant neighbors for which we had measured regulation strength. We observed only a very weak correlation between connectivity and regulation strength (R=-0.1, -0.1, 0.01 for the CRP, Fis, and IHF landscapes, Figures S13-S15). Similarly, the relative connectivity of peak genotypes is only weakly correlated with their regulation strength (R=-0.05, -0.04, 0.06 for the CRP, Fis, and IHF landscapes). (Results, lines 321–330), indicating that strongly regulating genotypes are not preferentially oversampled or undersampled (Results, lines 321–330).

The second, and most important, analysis directly addresses the reviewer’s concern that experimental uncertainty could affect peak classification and, consequently, landscape navigability. We explicitly incorporated experimentally measured, genotype-specific noise estimates from biological replicates when comparing fitness values between neighboring genotypes. Using these uncertainty-aware comparisons, we then recomputed adaptive-walk dynamics and genotype visitation frequencies on the resulting noisy landscapes.

We observe strong correlations between visitation frequencies in the noise-free and noisy landscapes across all three transcription factors (new Supplementary Figure S35), indicating that evolutionary accessibility patterns are robust to realistic levels of experimental uncertainty. These analyses are described in the revised Results (lines 622–636) and in a new Supplementary Methods section (“Incorporation of experimental uncertainty into adaptive walks”).

Reviewer #2 (Public review):

The authors aim to investigate the ability of evolution to create strong transcription factor binding sites (TFBSs) de novo in E. coli. They focus on three global transcriptional regulators: CRP, Fis, and IHF, using a massively parallel reporter assay to evaluate the regulatory effects of over 30,000 TFBS variants. By analyzing the resulting genotype-phenotype landscapes, they explore the ruggedness, accessibility, and evolutionary dynamics of regulatory landscapes, providing insights into the evolutionary feasibility of strong gene regulation. Their experiments show that de novo adaptive evolution of new gene regulation is feasible. It is also subject to a blend of chance, historical contingency, and evolutionary biases that favor some peaks and evolutionary paths.

(1) Strengths of the methods and results:

The authors successfully employed a well-designed sort-seq assay combined with high-throughput sequencing to map regulatory landscapes. The experimental design ensures reliable measurement of regulation strengths. Their system accounts for gene expression noise and normalizes measurements using appropriate controls.

Comprehensive Landscape Mapping:

The study examines ~30,000 TFBS variants per transcription factor, providing statistically robust and thorough maps of the regulatory landscapes for CRP, Fis, and IHF. The landscapes are rigorously analyzed for ruggedness (e.g., number of peaks) and epistasis, revealing parallels with theoretical uncorrelated random landscapes.

Evolutionary Dynamics Simulations:

Through simulations of adaptive walks under varying population dynamics, the authors demonstrate that high peaks in regulatory landscapes are accessible despite ruggedness. They identify key evolutionary phenomena, such as contingency (multiple paths to peaks) and biases toward specific evolutionary outcomes.

Biological Relevance and Novelty:

The author's work is novel in focusing on global regulators, which differ from previously studied local regulators (e.g., TetR). They provide compelling evidence that rugged landscapes are navigable, facilitating de novo evolution of regulatory interactions. The comparison of landscapes for CRP, Fis, and IHF underscores shared topographical features, suggesting general principles of global transcriptional regulation in bacteria.

(2) Weaknesses of the methods and results:

Undersampling of Genotype Space:

While the quality filtering of the data ensures robustness, ~40% of the TFBS space remains uncharacterized. The authors acknowledge this limitation but could improve the analysis by employing subsampling or predictive modeling.

We thank the reviewer for raising this point. We agree that undersampling of genotype space is an important limitation of our dataset and that, in principle, subsampling or predictive modeling approaches could be used to address missing genotypes. We have now clarified in the manuscript why these approaches are not straightforward in the context of our analyses and why we did not pursue them here.

Although approximately 40% of TFBS genotypes were removed during the filtering step due to lack of reliable measurements, this filtering step was necessary to ensure robust estimation of regulation strength from sort-seq data. Importantly, random subsampling of the genotypes in our data set would not alleviate this limitation, because many of our key analyses—such as peak identification, quantification of epistasis, and assessment of evolutionary accessibility—require combinatorially complete local neighborhoods in genotype space. Subsampling would remove mutational neighbors from many neighborhoods, and thus further limit our ability to characterize landscape topology.

Predictive modeling approaches could, in principle, be used to infer missing genotypes and reconstruct more complete landscapes. However, developing, experimentally validating, and benchmarking such models would not only substantially expand the scope of an already long paper, it would also require additional assumptions about genotype–phenotype relationships that entail their own limitations. Our primary goal in this work was to provide the first large-scale empirical in vivo regulatory landscapes for global bacterial transcription factors, comprising tens of thousands of experimentally measured variants. We view these empirical landscapes as a necessary foundation upon which predictive modeling and landscape completion can be built in future, complementary studies.

We have now revised the Discussion (lines 760-770) to explicitly articulate these points and to clarify that, while undersampling remains a limitation, it does not invalidate the landscape-level conclusions we draw from the combinatorially complete neighborhoods present in our data. There we also outline predictive modeling as an important directions for future work.

For a more detailed answer regarding subsampling and peak classification, please also see our response to comment (2) of Reviewer #1.

Simplified Regulatory Architecture:

The study considers a minimal system of a single TFBS upstream of a reporter gene. While this may have been necessary for clarity, this simplification may not reflect the combinatorial complexity of transcriptional regulation in vivo.

Point well taken. We have added paragraph to state explicitly that the system we use to study gene regulation is much simpler than most in vivo regulatory circuits (Discussion, lines 797-802)

Lack of Experimental Validation of Simulations:

The adaptive walks are based on simulated dynamics rather than experimental evolution. Incorporating in vivo experimental evolution studies would strengthen the conclusions. Although this is a large request for the paper, that would not prevent publication.

We thank the reviewer for this important point. We fully agree that in vivo experimental evolution would provide a valuable and complementary way to validate the evolutionary dynamics inferred from our simulations. However, we ask for the reviewer's understanding that adding experimental evolution to an (already long) paper would go far beyond the scope of our study.

Also, the goal of our study was not to reproduce evolutionary trajectories experimentally, but to characterize the structure of large empirical regulatory landscapes, and to use these landscapes as a data-driven basis for exploring evolutionary accessibility under well-defined population-genetic assumptions. The adaptive walks we employ are parameterized directly from experimentally measured genotype–phenotype maps, and incorporate established fixation probabilities. Such walks have been widely used to study evolutionary dynamics on empirical landscapes when experimental evolution is not tractable, because it would involve tens of thousands of genotypes that represent small mutational targets and would thus take a long time to evolve.

An additional issue related to the feasibility of experimental evolution is that performing in vivo experimental evolution for the regulatory landscapes analyzed here would require tracking large populations across a combinatorially vast TFBS space, while simultaneously measuring regulatory phenotypes for thousands of evolving lineages, which is currently not experimentally feasible. This is another reason why simulation-based approaches have been the standard method for linking large-scale empirical landscapes to evolutionary dynamics in both theoretical and experimental studies.

Furthermore, our conclusions are intentionally framed at the level of statistical and landscape-wide properties (e.g., accessibility of high peaks, contingency, and evolutionary bias), rather than at the level of specific mutational trajectories. As such, they do not rely on the precise reproduction of any single evolutionary path, but on aggregate patterns that are robust to reasonable variation in population-genetic parameters.

In sum, we do not view experimental evolution as essential for the conclusions we draw, but as an important and exciting direction for future work that may be enabled by the landscapes we have experimentally mapped.

Impact on the Field:

This study advances our understanding of adaptive landscapes in gene regulation and offers a critical step toward deciphering how global regulators evolve de novo binding sites. The findings provide foundational insights for synthetic biology, evolutionary genetics, and systems biology by highlighting the evolutionary accessibility of strong regulation in bacteria.

Utility of Methods and Dat

The sort-seq approach, combined with landscape analysis, provides a robust framework that can be extended to other transcription factors and systems. If made publicly available, the study's data and code would be valuable for researchers modeling transcriptional regulation or studying evolutionary dynamics.

Additional Context:

The study builds on a growing body of work exploring regulatory evolution. For instance, recent studies on local regulators like TetR and AraC have revealed high ruggedness and epistasis in TFBS landscapes. This study distinguishes itself by focusing on global regulators, which are more biologically complex and influential in bacterial gene networks. The observed evolutionary contingency aligns with findings in other biological systems, such as protein evolution and RNA folding landscapes, underscoring the generality of these evolutionary principles.

Conclusion:

The authors successfully mapped the genotype-phenotype landscapes for three global regulators and simulated evolutionary dynamics to assess the feasibility of strong TFBS evolution. They convincingly demonstrate that ruggedness and epistasis, while prominent, do not preclude the evolution of strong regulation. Their results support the notion that gene regulation evolves through a blend of chance, contingency, and evolutionary biases.

This paper makes a significant contribution to the understanding of regulatory evolution in bacteria. While minor limitations exist, the authors' methods are robust, and their findings are well-supported. The work will likely be of broad interest to researchers in molecular evolution, synthetic biology, and gene regulation.

We thank the reviewer for their thorough evaluation and for their supportive opinion of this paper.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Line 28 (Abstract): "Landscape ruggedness does not prevent the evolution of strong regulation, because more than 10% of evolving populations can attain one of the highest peaks." I did not find this interpretation very convincing; only 10% of populations being able to achieve strong regulation sounds to me like ruggedness DOES impede adaptation in the vast majority of cases.

We thank the reviewer for this thoughtful comment and agree that our original phrasing in the Abstract overstated this conclusion. We did not intend to imply that landscape ruggedness has only a minor effect on adaptation. On the contrary, our results clearly show that ruggedness strongly constrains evolutionary outcomes and prevents the majority of evolving populations from reaching the globally highest regulatory peaks. We have therefore toned down the wording in both the Abstract and the Discussion (lines 670-679) to reflect this more accurately. For example, in the abstract we now state

“Nonetheless, evolutionary simulations show that ~10% of evolving populations can reach a peak of strong regulation, a proportion that is significantly greater than in comparable random landscapes.”

In the discussion we state:

“… Specifically, our evolutionary simulations show that 10% of populations with a size typical of E. coli reach one of the highest peaks. This percentage is significantly higher than in randomized landscapes (Supplementary Methods 9; Supplementary Figure S30)"

Our intended interpretation was more limited: namely, that ruggedness does not fully preclude the evolution of strong regulation. In highly rugged landscapes with extensive sign epistasis—whose topological properties approach those of uncorrelated random landscapes—the a priori expectation is that access to the strongest peaks could be vanishingly rare or effectively impossible under Darwinian evolution. In this context, observing that a non-negligible fraction of populations (on the order of 10%) can reach one of the highest peaks suggests that strong regulation remains evolutionarily attainable, even though it is far from guaranteed.

Motivated by the reviewer’s suggestion, we also added a null-model analysis that makes this point more explicitly and quantitatively. Specifically, we constructed randomized landscapes by permuting regulation-strength values across genotypes while preserving the experimentally sampled genotype network topology and all parameters of the evolutionary simulations (Supplementary Methods 9, “Randomized landscape null model for peak accessibility”). We then repeated the adaptive-walk simulations on these shuffled landscapes. This null model provides an expectation for peak accessibility in landscapes with identical sampling, neighborhood structure, and evolutionary dynamics, but without genotype–phenotype correlations.

Using this null model, we find that the fraction of populations that reach high peaks in the empirical landscapes is substantially higher than expected by chance alone (new Supplementary Figure S30; Results, lines 504–516). Specifically, across the three transcription factors, empirical landscapes exhibit on average a ~3-fold higher accessibility of high regulatory peaks than shuffled landscapes. This comparison does not weaken the conclusion that ruggedness strongly impedes adaptation; rather, it shows that the structure of the measured genotype–phenotype landscapes enables greater accessibility of strong regulation than would be expected in equally rugged but unstructured landscapes.

In response to the reviewer’s concern, we have revised the abstract and main text to avoid the phrase “does not prevent” and to more accurately convey this balance between constraint and accessibility. We now emphasize that ruggedness strongly constrains adaptation, while still allowing access to strong regulatory peaks at rates that exceed null expectations. (Discussion, lines 512-516). For example, in the discussion we state:

“… In sum, rugged regulatory landscapes strongly constrain evolutionary trajectories, yet do not render the evolution of strong regulation vanishingly rare. Instead, strong regulatory phenotypes remain evolutionarily attainable at levels that exceed null expectations, even though they are reached by only a minority of evolving populations.”

We believe that the revised wording, together with the added null-model analysis more faithfully represents our results and strengthens the quantitative interpretation of accessibility in these landscapes.

(2) Line 123: I found the explanation of the plasmid system and the accompanying SI figures (Figures S1 and S2) confusing in terms of how many plasmids there were. In particular, the Figure S1 graphics show the plasmid specifically with CRP but the text in the graphic and in the caption refers to the plasmid pCAW-Sort-Seq-V2 (which, according to Table S1, isn't that just the base plasmid without any TF?). Figure S2 also shows the plasmid with CRP and does specify pCAW-Sort-Seq-V2-CRP-CRP0 in the graphic, but then the caption refers again only to the base plasmid pCAW-Sort-Seq-V2. I recommend the authors clarify these items for readers who might want to reproduce or build upon their system. In particular, I recommend the main text explain more explicitly that they generate three versions of this plasmid (one for each TF), and then on the backgrounds of each of those three plasmids, a whole library with all the binding site variants.

We thank the reviewer for pointing out this lack of clarity. We agree that the original description of the plasmid system and the accompanying Supplementary Figures S1 and S2 could be confusing with respect to how many plasmids were used and how they differ.

To clarify the experimental design, we start from a common backbone plasmid, pCAW-Sort-Seq-V2, which contains all shared regulatory and reporter elements but does not encode any transcription factor. From this backbone, we generated three distinct TF-specific plasmids, each carrying one of the transcription factors studied here—CRP, Fis, or IHF—resulting in pCAW-Sort-Seq-V2-CRP, pCAW-Sort-Seq-V2-Fis, and pCAW-Sort-Seq-V2-IHF. On the background of each TF-specific plasmid, we then constructed a complete library of plasmids containing all variants of the corresponding TF binding site cloned upstream of the reporter gene.

We have revised the main text to explicitly describe this plasmid hierarchy and library construction strategy and to clarify that three TF-specific plasmids were generated prior to TFBS library construction (Results, Landscape mapping section; lines 159–193). In addition, we have redesigned Supplementary Figures S1 and S2 to facilitate understanding of the plasmid system. Specifically, these figures now clearly distinguish between the base plasmid backbone and the TF-specific plasmid derivatives. Also, the plasmid names shown in the graphics and captions are now consistent with those listed in Supplementary Table S1. Upon final publication, we will also deposit the sequences of all plasmids in Addgene to further facilitate reproducibility.

(3) Line 135: Can the authors clarify whether these TFs are essential in these media conditions and, if not, why? I was expecting them to be so given the core functions of these TFs as described in the Introduction, but then Figure S3 appears to show that all knockouts are viable.

We thank the reviewer for raising this important point and apologize for the lack of clarity in the original version of the manuscript. The transcription factors CRP, Fis, and IHF are not essential for viability under the growth conditions used in this study, but they are important for optimal growth and cellular fitness, consistent with their roles as global regulators.

Under our experimental conditions, single-gene knockout strains (Δcrp, Δfis, and Δihf) are viable but exhibit slower growth dynamics compared to the wild-type strain, reflecting impaired regulation of core cellular processes (Supplementary Figure S3). This behavior is consistent with previous work showing that many global transcriptional regulators in E. coli are conditionally essential or strongly fitness-affecting, rather than absolutely essential under standard laboratory conditions.

Importantly, while single knockouts remain viable, double mutants involving these global regulators are not viable, indicating substantial functional redundancy and network-level essentiality among global transcription factors. This explains why each TF can be studied individually in isolation, while combinations of deletions cannot be maintained.

We have now clarified this point in the Results section by explicitly stating that the knockout strains show reduced growth rates but reach comparable cell densities during late exponential or early stationary phase, the growth phase at which all measurements were performed (Results, Landscape mapping section; lines 185–193). This clarification reconciles the apparent discrepancy between the biological importance of these transcription factors discussed in the Introduction and the viability of the single-knockout strains shown in Supplementary Figure S3.

(4) Lines 141 and 227: The authors appear to refer to two different citations for different versions of RegulonDB (refs. 47 and 66). Did they actually use both versions for different purposes (if so, why?), or is this a typo?

We thank the reviewer for noticing this inconsistency. We did not use two different versions of RegulonDB. The two separate references were an error. We have now corrected this by using a single, consistent RegulonDB citation in both locations.

(5) Line 166 (Figure 1 caption): I think 2^8 here should be 4^8.

Thank you. We have corrected “28” to “48” in the Figure 1 caption.

(6) Figure 2Are the distributions in Figure 2a (regulation strengths across all TFBSs in the libraries) equivalent to the distributions in Figures S4-S6 (direct fluorescence readout from cell sorting), just transformed from fluorescence to regulation strength? If so I think that would be helpful to clarify, perhaps in the captions to Figures S4-S6 so that it's clear these contain the same information.

No. Figures S4–S6 and Figure 2a do not show the same distributions. Figures S4–S6 display the raw fluorescence distributions obtained from cell sorting, whereas Figure 2a shows regulation strengths (S), which are derived quantities computed from these fluorescence data. Specifically, regulation strength is calculated as a weighted average over fluorescence bins using the sequencing read distribution for each TFBS (see Methods, “Regulation strengths”).

To clarify this relationship, we have revised the main text (lines 201-203 and Figure 1b-c), to explicitly state how regulation strengths (S) were calculated.

(7) Figure 2b: Can the authors label each logo/frequency matrix with its corresponding TF name in the graphic itself? I think this is only implied in the caption.

We have updated Figure 2b to label each sequence logo / frequency matrix directly in the graphic with its corresponding transcription factor name (CRP, Fis, or IHF), in addition to mentioning these names in the caption. This change clarifies the figure and makes the TF identity immediately apparent to the reader.

(8) Lines 290 and 298 (Figure 2 caption): The labels for panels b and c appear to be swapped in the caption.

We thank the reviewer for pointing this out. The labels for panels b and c in the Figure 2 caption were indeed swapped. This has now been corrected.

(9) Line 379: There is a missing period at the end of this line.

We have added the missing period at the end of this line.

(10) Line 400 (Figure 3 caption): There is a missing subtitle for panel c in the caption for this figure (all other panels seem to have bolded subtitles in their captions).

We have added the missing subtitle for panel c in the Figure 3 caption to match the formatting of the other panels.

(11) Line 583: There is a missing period after "Methods 7.5)".

We have added the missing period after “Methods 7.5)”.

(12) Line 641: "All three landscapes highly rugged" should probably be "All three landscapes are highly rugged".

We have corrected the sentence to read “All three landscapes are highly rugged.”

Version 2

Abstract

The evolution of gene regulation is a major source of evolutionary adaptation and innovation, particularly when organisms encounter new or changing environments. Central to this process is the emergence of new transcription factor binding sites (TFBSs). Adaptive landscapes provide a powerful framework to study such emergence by linking regulatory DNA sequences to their transcriptional outputs. Although several landscapes have been characterized for DNA, RNA, and proteins, large-scale in vivo adaptive landscapes for bacterial TFBSs remain scarce. Here, we address this gap by experimentally mapping the first comprehensive in vivo regulatory landscapes for three global transcription factors in Escherichia coli: CRP, Fis, and IHF. Using a massively parallel reporter assay, we quantify the regulation strength of more than 30,000 TFBS variants for each factor and reconstruct their adaptive landscapes. All three landscapes are highly rugged and exhibit pervasive epistasis, with thousands of local peaks distributed broadly across sequence space. This ruggedness contrasts sharply with the much smoother TFBS landscapes of eukaryotes. It suggests greater constraints on the evolution of prokaryotic gene regulation. Nonetheless, evolutionary simulations show that ∼10% of evolving populations can reach a peak of strong regulation, a proportion that is significantly greater than in comparable random landscapes. Adaptive evolution starting from the same DNA sequence can attain different high peaks, and some peaks are reached more frequently than others. Together, our results show that de novo adaptive evolution of new gene regulation in bacteria is feasible, but subject to a blend of chance, historical contingency, and evolutionary biases.

Introduction

Transcriptional regulation controls the ability of RNA polymerase to initiate the transcription of a gene into mRNA1. It is crucial in the life of all organisms, orchestrating the expression of genes in response to environmental cues and cellular states2,3. Transcriptional regulation is mediated by transcription factors (TFs), proteins that bind DNA near a gene at short DNA words known as transcription factor binding sites (TFBSs). The binding of a TF to its binding site on DNA can either hinder or facilitate transcription initiation by RNA polymerase1,4,5. The stronger the binding of a TF to its TFBS is, the more strongly the TF can regulate a nearby gene6–9. In bacteria, TFs operate within a hierarchically organized gene regulatory network. At the bottom of this hierarchy are local TFs that regulate the expression of one or few genes and modulate specific biological processes 10–12. At the top are global TFs that may regulate hundreds of genes and control many cellular functions 11,13–15. Consistent with their broader role, global regulators are typically more highly expressed and bind to a broader range of TFBSs than local regulators12.

A genotype-phenotype map is a conceptual analog of a physical landscape, where each location corresponds to one genotype in sequence space, and is associated with a quantitative phenotype. If the phenotype is related to gene expression, the map is also called a regulatory landscape6,16–19. Another special case of such a map is a fitness landscape or adaptive landscape, in which the phenotype is fitness and is interpreted as an elevation20,21. Over the last decade, genotype–phenotype (GP) maps and fitness landscapes have become central tools for understanding how molecular systems evolve under mutation and selection22–25. Such maps and landscapes have been experimentally studied for DNA6,8,18,19,26,27, protein28–32 and RNA33–35 molecules, revealing key topographical properties that shape evolutionary outcomes, including epistasis24,36—the non-additive effects of multiple mutations on phenotype—landscape ruggedness, reflected in the number and distribution of fitness peaks, and constraints on adaptive evolution. For example, one large-scale study adopted CRISPR-Cas9 technology to measure the fitness of more than 200,000 E. coli genotypes that encode variants of the bacterial antibiotic resistance gene dihydrofolate reductase (DHFR). It showed that this landscape is highly rugged (multi-peaked)31.

For transcription factor binding sites, most pertinent large-scale studies are based on in vitro binding assays, such as protein-binding microarrays (PBMs), and they focus predominantly on eukaryotic transcription factors6. While these studies have been instrumental in characterizing transcription factor binding preferences, they typically do not measure regulatory output in a native cellular context. In contrast, comprehensive in vivo data for bacterial TFBSs remain extremely rare. To our knowledge, only two high-resolution in vivo landscapes have been previously mapped for bacterial regulators, those of the local regulators TetR18 and LacI27. As a result, it remains unclear whether principles inferred from protein landscapes, eukaryotic TFBSs, or in vitro binding assays generalize to transcriptional regulation in bacteria, particularly for global regulators11 that integrate multiple physiological signals.

Both TFs and their TFBSs evolve, but TFBSs evolve more rapidly. The reason is that a mutation in a TF can affect the expression of many genes, whereas a mutation in a TFBS may affect only one gene and is thus less likely to be deleterious 13,37–39. A special case of TFBS evolution is the evolution of a strong TFBS from a DNA word with weak or no regulatory activity. Such de novo evolution of TFBSs can create new regulatory interactions and change the structure of gene regulatory networks40–43. Unfortunately, we know little about the ability of Darwinian evolution to create TFBSs de novo. A strong TFBS may have to arise from a non-binding site through an evolutionary path of multiple mutational steps. Darwinian evolution can favor this process only if strong binding is adaptive and if each mutational step in a path increases binding strength, i.e., if the path is accessible to Darwinian evolution. To find whether such paths exist may require the analysis of multiple paths. This is challenging because sequence space contains an astronomical number of potential TFBSs for any one TF. The number of evolutionary paths to strong transcriptional regulation is thus also astronomical. For each such path, the strength of each TFBS along the path has to be measured experimentally 44–47.

In principle, one could attempt to construct such landscapes in silico using commonly employed models of TF–DNA interactions, such as position weight matrices (PWMs)48–50. However, PWMs are derived from a limited set of naturally occurring binding sites and primarily reflect sequence conservation rather than quantitative regulatory output. Moreover, PWMs assume independent and additive contributions of individual nucleotide positions to DNA binding 48–50. They therefore cannot capture the influences of epistatic interactions between positions, which can dramatically affect landscape topography and evolutionary accessibility24. Lastly, PWMs do not account for important biological effects that modulate gene regulation such as DNA shape51,52, cooperative interactions53,54, and chromosomal context55,56. Thus, experiments are necessary not only to obtain quantitative measurements of gene regulatory activity, but also to refine and inform PWM-based models using thousands of experimentally characterized sequences.

Building on our previous work on a local TF18, we address this challenge for three global regulators in Escherichia coli by performing three independent and massively parallel experiments44 for each TF. The experiments use a synthetic biology platform to quantify how strongly each of more than 30,000 binding sites for a TF can regulate the expression of a nearby reporter gene.

The first TF we study is the cAMP receptor protein (CRP). In the absence of glucose, CRP modulates the expression of genes mostly involved in carbon metabolism. It allows E. coli to efficiently switch between sources of carbon and energy 57–60. The second TF is the factor for inversion stimulation (Fis), which helps to regulate the expression of genes involved in growth phase transitions. It also modulates the supercoiling of DNA61–63 and influences DNA replication, recombination, and repair61–63. The third factor is the integration host factor (IHF). IHF regulates genes involved in stress responses and stationary phase survival64,65. Similar to Fis, it is involved in DNA compaction, replication, and recombination, but also in the assembly of complex nucleoprotein structures64,65. We chose these factors because they are the most global regulators in E. coli, and they are diverse, belonging to different protein families.

We use our experimental data for each TF’s binding sites to map the relationship between binding site genotype and gene expression. Our first aim is to characterize the resulting regulatory landscapes for global bacterial regulators, and to find out whether these landscapes are different or similar. When strong regulation of a gene is adaptive, a regulatory landscape becomes a fitness landscape 20,21,25,66. Our second aim is to understand how populations would evolve on each of our three landscapes when they are viewed as fitness landscapes. Specifically, we study how evolving populations would traverse each landscape through individual mutational steps that change a TFBS its ability to regulate a gene via its cognate TF. A peak in such a landscape is a TFBS conveying stronger regulation than all neighboring TFBSs in sequence space. If such a landscape is rugged (has multiple peaks), natural selection alone may not enable a population to discover the highest peaks, i.e., the strongest TFBSs. The reason is that the peaks may be separated by valleys of low regulation strength that cannot be traversed by natural selection alone 67–69 25. In other words, high peaks may not be easily accessible through Darwinian evolution – they may be reachable by few or no evolutionary paths of single DNA mutations in which each mutational step increases TFBS strength6. One factor that can reduce peak accessibility is epistasis, which can reduce the predictability of evolutionary trajectories towards a peak 24,36.

To accomplish both aims, we first studied the topography of the three landscapes, and then simulated the dynamics of evolving populations on them. All three landscapes are highly rugged. They contain more than 2,000 peaks that are scattered through genotype space and are rife with epistatic interactions, in striking contrast to the comparatively smooth TFBS landscapes described for eukaryotic systems6. Despite these features, evolving populations can reach the strongest TFBSs in all three landscapes. Which of several high peak is reached is contingent on chance events during adaptive evolution.

Results

Landscape mapping

We constructed a modular plasmid system based on the common backbone plasmid pCAW-Sort-Seq-V218 (Supplementary Table S1; Supplementary Figures S1–S2; Supplementary Methods 2–3). This backbone contains all shared regulatory and reporter elements required for fluorescence-based measurements, but it encodes neither a TF nor its binding site(s) for any of the regulators studied here. From this backbone, we generated three TF-specific plasmid derivatives. Each of these plasmids encodes one of the global transcription factors CRP, Fis, or IHF under inducible control (hereafter pCAW-Sort-Seq-V2-CRP, pCAW-Sort-Seq-V2-Fis, and pCAW-Sort-Seq-V2-IHF; Supplementary Figure S2). In each of the TF-specific plasmids, a TFBS insertion site is positioned immediately upstream of the gfp reporter gene, such that TF binding represses the transcription of gfp. Consequently, stronger TF–DNA binding results in lower GFP expression, which enables a quantitative readout of a binding site’s regulation strength.

For each TF-specific plasmid, we constructed three kinds of variants. The first carries a wild-type (WT) TFBS for the corresponding TF upstream of gfp. It serves as a reference conferring wild-type regulation of the reporter gene. The second contains a complete TFBS library, in which we randomized the eight most information-rich base-pair positions of the respective binding site (Supplementary Methods 4; Supplementary Table S3), as determined from alignments of experimentally characterized binding sites curated in RegulonDB70. Each library comprised 4⁸ = 65,536 unique TFBS sequences, and we constructed three independent biological replicate plasmid libraries from them per TF (Supplementary Methods 4–5). The third variant lacks a promoter upstream of gfp and serves as a negative control that allows us to quantify cellular autofluorescence during fluorescence measurements (Supplementary Table S1).

We introduced each TF-specific plasmid into an E. coli host strain in which the corresponding chromosomal TF gene had been deleted (Δcrp, Δfis, or Δihf; Supplementary Table S2). This genetic background ensures that the focal transcription factor is expressed exclusively from the plasmid. Although the mutant strains grow more slowly than the WT, they reached similar cell densities during late exponential or early stationary phase, the growth phase at which we performed all measurements (Supplementary Figure S3). TF expression is controlled by a tetracycline-inducible promoter and can be precisely tuned using anhydrotetracycline (aTc), allowing us to regulate TF abundance independently of growth conditions and to isolate the effects of TF–DNA binding on transcriptional regulation (Figure 1a).

Mapping TFBS regulatory landscapes.

a-c) Sort-Seq procedure. We utilized a plasmid-based fluorescence reporter system followed by sort-seq to map TFBS regulatory landscapes. a. Library generation. For generating our E. coli libraries, we cloned TFBS sequence variants (48 = 65,536), represented as black circles, into our plasmid, between a σ70 constitutive promoter (shown as a black right-facing arrow) and a gfp gene (shown as a green right-facing arrow), to measure transcriptional repression through fluorescence intensity. When a TF binds to the TFBS, it blocks RNA polymerase activity through steric hindrance, reducing gfp transcription. Thus, lower fluorescence levels (light green) represent stronger TF-TFBS binding. The TFBS variants produce a range of regulation strengths resulting in variable green fluorescence intensities in bacterial cells (green-colored rounded rectangles). b. Sorting procedure. We sorted libraries into expression bins based on fluorescence intensity using a fluorescence-activated cell sorting (FACS, Supplementary Figures S4-S6). c. Sequencing and phenotyping. We sequenced TFBS variants from each fluorescence bin and used these data to calculate a continuous regulation strength for each genotype. Regulation strength (S) was computed as a weighted average of fluorescence across bins, based on the distribution of sequencing reads (see Methods). Regulation strength is visualized using a color gradient from dark green (low-affinity TFBSs, high fluorescence) to light green (high-affinity TFBSs, low fluorescence). d) Genotype-phenotype mapping. To construct a regulatory landscape, we connected TFBS genotypes (colored circles) that differed by a single nucleotide via edges, thereby establishing an interconnected genotype network. Each genotype conveys a regulatory phenotype (strength of regulation, heatmap colors), which can be viewed as the elevation dimension (z-axis) in a landscape.

We then mapped these TFBS genotypes to their respective regulatory phenotypes using a well-established technique known as sort-seq16,44,71–73 (Figure 1b), which combines fluorescence-activated cell sorting (FACS) with high-throughput sequencing (Figure 1c). In sort-seq, one first sorts cells into multiple fluorescence “bins” depending on the level of GFP expression. A cell’s GFP fluorescence serves as a proxy for GFP expression levels74 and quantifies how strongly the TFBS library member in this cell can regulate GFP expression (Methods, Supplementary Methods 5). For each TFBS genotype, we quantified regulation strength (S) as a weighted average of its sequencing counts across the different fluorescence bins, yielding a single continuous measure of regulatory activity (see Methods).

The results of our sort-seq experiments are three maps – one for each transcription factor – from each of more than 30,000 genotypes (TFBS variants) to regulatory phenotypes (regulation strength). One can view each map as a regulatory landscape8,9,16,18,75–79 (Figure 1d) that becomes a fitness landscape whenever strong regulation entails high fitness6,25. For the purpose of analyzing the landscape quantitatively, we represent it as a network of genotypes (TFBS variants). Each node in this network corresponds to a genotype. Edges link neighboring genotypes, which differ in a single nucleotide (Figure 1d).

Landscapes exhibit diverse regulation strengths and distribution breadths

To evaluate the ability of our library to regulate gene expression, we first measured the distribution of GFP fluorescence intensities across the bins in two conditions, i.e., in the presence or absence of the TF (with or without the atc inducer, Supplementary figures S4-S6). In the presence of the TF, GFP expression was lower on average and showed a broader distribution than in the absence of the TF (Supplementary figures S4-S6). This indicates that the TFBSs in each library can indeed downregulate GFP expression, but some library variants convey stronger regulation than others, hence the broader fluorescence distribution.

We then pooled barcoded DNA sequences extracted from cells in each bin and biological replicate. We sequenced at least 250 unique TFBS genotypes from each bin, with an average of 3992.3±4293.9 unique sequences per bin for the three TFs (as detailed in Methods, Supplementary Methods 7, and Supplementary Figures S4-S6). The resulting sequences covered 95%, 90%, and 93% of the total library sizes (N = 65,536 genotypes) for CRP, Fis and IHF, respectively. To ensure the reliability of our data, we excluded sequences with low read coverage and sequences that were not present in all triplicates (Methods, Supplementary Methods 7.3). This quality filtering step resulted in library sizes of 31,975 genotypes for CRP (49% of all 48 genotypes), 43,222 genotypes for Fis (66%), and 41,325 genotypes for IHF (63%). The correlation in read counts for each variant across replicates was very high for this quality-filtered data (Pearson’s R= 0.98-0.99) (Supplementary Figures S7-S9). We used this data for all further analyses.

The fluorescence and sequence data from each bin allowed us to quantify the variant’s ability to regulate gene expression for each TFBS library variant. We refer to the resulting metric as the regulation strength S conveyed by the variant (Methods, Supplementary Methods 7.2, Figure 1d). It ranges between S=0 (no regulation) to S=1 (strongest regulation among all variants in the library). Although our experiments directly quantify expression regulation, and not the affinity or binding strength of a TF to a TFBS variant, we also use binding strength as a proxy for regulation strength, because TF-DNA binding is necessary for regulation1,4,11,54,80,81. To each of our three TFs, we also assigned a reference TFBS that is naturally occurring and conveys strong regulation by the TF, as proven by previous experimental work60,82–84. We refer to this TFBS as the wild-type (WT 82,85 WT 83,86 and WT 84,87, Supplementary Table S4). We refer to TFBSs with regulation strengths below and above the wild-type (WTCRP: SWT=0.71, WTFis: SWT=0.97, WTIHF: SWT=0.95) as weak and strong, respectively.

We observed a broad range of regulation strengths S for each TF landscape, with varying dispersions. The CRP landscape exhibited the lowest average regulation strength and a narrower distribution of S compared to Fis and IHF, which had similar distributions (mean±s.d., 0.37±0.1 for CRP, 0.57±0.13 for Fis, and 0.52±0.14 for IHF; see Figure 2a, Supplementary Figure S10). Next, we analyzed the strongly regulating TFBSs to identify nucleotides that may be particularly frequent, and thus potentially important for strong regulation (Figure 2b). We discovered a moderate association between the most frequent nucleotides (highlighted in yellow in Figure 2b) and the most informative nucleotides from the available position weight matrices (PWMs) for these TFs70. (Pearson correlation coefficients: R=0.51, R=0.43, R=0.47 for Fis, CRP and IHF, respectively, with p-values smaller than 10-16, rejecting the null hypothesis of an absence of association). This observation suggests that our logos capture different information compared to available PWMs. This is expected, because our approach allows us to filter sequences by regulation strength thresholds to construct our matrices, unlike the traditional method of aligning genomic TFBSs without considering their regulation strengths48,49.

a. Genotypes in each landscape vary broadly in their regulation strength. The violin plots show the distribution of regulation strengths S (vertical axis) for each TF landscape (horizontal axis). The width of a plot at a given value of S represents the frequency of TFBSs at this value. The vertical length of the box in each violin plot covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR. The white circle shows the regulation strength of the wild type for each landscape (CRP: 0.71, Fis: 0.97, IHF: 0.95). The landscape sizes are as follows. CRP: N=31,975 TFBS variants; Fis: N= 43,222; IHF: N= 41,325. b. Peak genotypes are stronger regulators than non-peak genotypes. Dual violin plots show the distribution of regulation strength (vertical axis) for the three TF landscapes (horizontal axis), stratified by non-peak genotypes (grey, CRP, N=29,821, Fis, N= 40,910, IHF, N= 38,872), and peak genotypes (red, CRP, N= 2,154, Fis, N= 2,312, IHF, N= 2,453). The black tick-mark in each plot indicates the mean regulation strength of both non-peak and peak genotypes taken together (mean ± standard deviation, CRP: 0.37 ±0.1, Fis: 0.57±0.13, IHF: 0.52±0.14). The white circle on each plot marks the regulation strength of the wild-type (CRP: 0.71, Fis: 0.97, IHF: 0.95). c. Sequence logos and nucleotide frequency matrices for strong CRP, Fis and IHF binding sites. Each sequence logo 48,49 is based on an alignment of TFBSs with greater regulation strength than the wild-type for each of the three TFs IHF, Fis, and CRP. Each logo also shows the non-varying position (grey) of the TFBS genotype from which our libraries were created. In each logo, the height of each letter at each TFBS position indicates the information content at that nucleotide position – the taller the letter, the more frequent the nucleotide is in strongly regulating TFBSs 48,49. Similar information is conveyed by the frequency matrices displayed as heat maps below each logo. They represent the variability of each nucleotide at each position (horizontal axis) through a color gradient (see color legend). Tall letters in the sequence logo and yellow letters in the frequency matrix indicate frequent, and thus likely important nucleotides for strong regulation.

To further validate the data from our sort-seq experiments, we isolated cells harboring 10 different TFBS variants from each of 13 bins of each library (i.e., 10×13=130 variants per library), and determined their regulation strength more directly by quantifying GFP expression with a microplate reader (Methods and Supplementary Figure S11). This comparison validates the sort seq approach by revealing a strong association of the two independent quantifications of regulation strength (Pearson’s R=-0.81 for CRP, R=-0.74 for Fis, and R=-0.73 for IHF). (Methods and Supplementary Figure S11).

All three regulatory landscapes are highly rugged

The study of our landscapes in a network framework (Figure 1d) can help to quantify different aspects of landscape topography (Supplementary Table S5). One of them is the ruggedness of each landscape. It can be quantified by the number of peaks31,88–90. In the network framework, a peak is a TFBS whose neighbors are all weaker regulators (with lower S) than the TFBS itself. We find that all three landscapes are highly rugged, with 2,154, 2,312, and 2,453 peaks for the CRP, Fis, and IHF landscapes, respectively (Figure 2c, Supplementary Figure S12, Supplementary Table S5). Not surprisingly, peak genotypes generally are stronger regulators than non-peak genotypes (Figure 2c). Only a small fraction of peak genotypes regulate expression more strongly than the wild-type (Figure 2c; 61 for CRP, 172 for Fis, and 199 for IHF). We refer to such peaks as high peaks and distinguish them from low peaks (S<SWT).

The prevalence of sign epistasis (Supplementary Table S5) supports the notion that our landscapes are indeed rugged (see Supplementary Methods 7.5 for further details on epistasis and its evolutionary consequences). Independent evidence for landscape ruggedness comes from comparing the ruggedness of our landscapes with that of a well-established theoretical model of uncorrelated random landscapes, in which each sequence is assigned a fitness at random from the same fitness distribution, and neighbors have uncorrelated fitness values90–92. Such landscapes are maximally rugged 90–92. We created 103 uncorrelated random landscapes for each of our three TF landscapes by randomly shuffling the measured regulation strengths among all genotypes (Supplementary Methods 7.6). The number of peaks in our TF landscapes lies within 93%, 96%, and 98% percent of that of a maximally rugged random landscape, which, on average [mean±s.d.], has 2,308±133, 2,405±89 and 2,373±97 peaks for CRP, Fis and IHF, respectively. This analysis underlines that our landscapes are indeed highly rugged.

Because we use only quality-filtered genotype data, our landscapes lack regulatory information for about 40% of the 48 genotypes. While we cannot exclude that this undersampling of genotypes has led to systematic biases in our determination of landscape ruggedness, we note that the sampling of landscape genotypes by our experiments was not strongly biased with respect to regulation strength. Specifically, when we analyzed the relative connectivity of genotypes in our landscapes – the fraction of each genotype’s 24 possible neighbors for which our experiments yielded regulatory data – we found that it is only weakly correlated with regulation strength (R=-0.1, -0.1, 0.01 for the CRP, Fis, and IHF landscapes, Figures S13-S15). Similarly, the relative connectivity of peak genotypes is only weakly correlated with their regulation strength (R=-0.05, -0.04, 0.06 for the CRP, Fis, and IHF landscapes).

Landscape peaks are widely scattered in genotype space

In a rugged adaptive landscape, reaching high fitness peaks can be challenging, because such peaks are separated from other genotypes by valleys of low fitness that cannot be traversed by natural selection alone25. If natural selection favors strong gene regulation, the ruggedness of our regulatory landscapes may thus present a challenge for adaptive evolution. To better understand the potential magnitude of this challenge, we next analyzed our landscapes’ topography in greater detail. We began by studying the distribution of peaks in genotype space. If a landscape’s highest peaks are widely scattered through genotype space, then they may be accessible via fitness-increasing evolutionary paths from diverse non-peak genotypes. This may facilitate adaptive evolution compared to a landscape where peaks are clustered in a small region of genotype space.

For each of our three landscapes, we determined the genetic distance between peaks, i.e., the minimum number of mutations needed to transition from one peak to another, regardless of their effect on regulation strength. We compared the distribution of these distances to the distribution of genetic distances for an equal number of randomly selected non-peak variants. In all three landscapes, the distances between peaks are almost indistinguishable from those of random genotypes, differing on average by fewer than 0.1 mutational steps. In other words, peaks are about as widely dispersed in each landscape as random genotypes (Supplementary Figure S16). This is the case for both low and high peaks (Supplementary Figures S17-S18). A principal component analysis further underscores this dispersion (Supplementary Methods 7.7, Supplementary Figures S19-S21).

Accessible paths to a peak are often not the shortest possible paths

Next, we focused on mutational paths to peaks that are evolutionarily accessible, i.e., a series of mutational steps where each step increases regulation strength. Specifically, we enumerated all accessible paths that exist from each non-peak genotype to each peak genotype for all three landscapes. We found that these paths are generally longer than the shortest genetic distance between a non-peak and the peak genotypes. Specifically, the mean length of accessible paths exceeded the shortest distance by 1.5, 1.8 and 1.8 steps for the CRP, Fis and IHF landscapes (Supplementary Figure S22). In other words, accessible mutational paths are often not the most direct paths.

The existence of indirect paths implies that some mutational steps are evolutionarily prohibited because they decrease regulation strength. The reason is closely linked to non-additive (epistatic) effects of two or more mutations on regulation strength. (see Supplementary Methods 7.5 for further details on epistasis). More specifically, such inaccessible steps are a result of sign epistasis. In this kind of epistasis, a double mutant of a TFBS regulates expression more strongly than the TFBS itself, even though one or both constituent single mutants regulate expression more weakly than the TFBS24,93–95. Indeed, epistasis is prevalent in all three landscapes (Table S5). Specifically, we observe sign epistasis in 62%, 66%, and 65% percent of interactions between single mutant pairs in the CRP, Fis, and IHF landscapes.

The existence of accessible paths alone does not tell us how easily a high peak can be found through Darwinian evolution. The reason is that only a very small fraction of all paths to that peak may be accessible, and an evolving population may not find any one of those paths. To quantify the fraction of accessible paths, we determined the total number of paths from each non-peak genotype to each high fitness peak, and computed the fraction of those paths that are accessible. Figure 3a shows how this fraction depends on the number of mutational steps in a path. It behaves similarly for all three landscapes (Figure 3a). The majority of two-step paths (a fraction greater than 80%) are accessible in all three landscapes (Figure 3a), but the fraction of accessible paths dwindles rapidly with an increasing number of mutational steps. It reaches a minimum below 1% for all three landscapes at eight mutational steps (Figure 3a).

Peaks and their basins of attractions.

a. The fraction of accessible paths declines with increasing path length. The figure illustrates how the fraction of accessible paths to high peaks (vertical axis, logarithmic scale) decreases with the length of the shortest accessible path (horizontal axis). Accessible paths are defined as paths where each step increases regulation, starting from a specified initial genotype. Each colored line corresponds to data from a different TF landscape: CRP (green), Fis (orange), and IHF (blue). Circles indicate the mean fraction of accessible paths for a given path length. b. Higher peaks have larger basins of attraction. The split-violin plots display the distribution of basin sizes (vertical axis) for high (red) and low (blue) peaks in the three TF landscapes CRP, Fis, and IHF (horizontal axis). High peaks have significantly larger basins of attraction (Welch Two Sample t-tests; CRP: t-value = 9.0898, df = 63.509, and p-value = 4.22 × 10-13, with mean basin sizes of xhigh = 9528.426 and xlow = 5655.035. Fis: t-value = 7.617, df = 172.35, and p-value = 1.645 × 10-12, with mean basin sizes of xhigh = 14206.70 and xlow = 10931.94. IHF: t-value = 6.1777, df = 201.99, and p-value = 3.521 × 10-9, with mean basin sizes of xhigh = 10965.872 and xlow = 8664.953.). The violin plots show the distribution of basin sizes for each of the two kinds of peaks. Their width represents the frequency of a given basin size. The vertical length of the box in each plot covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR. c. Peak genotypes with larger basins of attraction regulate expression more strongly. The regulation strength of peak genotypes (horizontal axis) is plotted against the size of their basins of attraction (vertical axis, shown as a percentage of all (non-peak genotypes) for peaks in all three landscapes (color legend). Color-coded values of R at the top of the graph indicate Pearson correlation coefficients for each landscape (color legend). Curved lines are based on a linear regression analysis (in linear space), with grey shading indicating 95% confidence intervals (CRP: R2=0.28, N = 2,154; Fis: R2=0.24, N = 2,312; IHF: R2=0.23, N = 2,434). d. Basins of attractions share many TFBS genotypes. The violin plots with embedded boxplots illustrate the fraction of genotypes shared between basins of attractions (vertical axis) for all pairs of high peaks in the CRP (green), Fis (orange), and IHF (purple) TF landscapes (horizontal axis). Specifically, we computed basin overlap as one minus the Jaccard coefficient (Supplementary Methods 7.5). A value of one (zero) indicates that the basin of attraction of two peaks share all (no) genotypes (CRP: 61 high peaks, N=1,830 comparisons, Fis: 172 high peaks, N=14,706 comparisons, IHF: 199 high peaks, N=19,701 comparisons). Their width represents the frequency of a given basin overlap. The vertical length of the box in each box covers the range between the first and third quartiles (IQR). The horizontal line within the box represents the median value, and whiskers span 1.5 times the IQR.

High peaks have large basins of attraction that share many TFBS genotypes

In the following analysis, we computed the basin of attraction of each peak, defined as the set of non-peak genotypes from which accessible paths to the peak exist31,96,97. In other words, a peak’s basin of attraction comprises all genotypes from which adaptive evolution can access the peak. The peaks in our landscapes have widely different basin sizes. They include peaks accessible from a mere fraction of genotypes to those accessible by a majority of genotypes (Figure 3b). Notably, high peaks generally had larger basins of attraction than low peaks (Figure 3c, Supplementary Figure S23). In addition, we found that many genotypes are part of multiple basins of attraction. Adaptive evolution may reach different high peaks starting from any such genotype. For all pairs of high peaks, we computed the pairwise overlap (intersection) of the basins of attraction, i.e., the fraction of genotypes that are part of both basins of attraction. (Figure 3d, Supplementary Figures S24-S25). The number of genotypes in this intersection varies widely, but the basins of attraction of high peaks generally share a substantial fraction of genotypes (mean shared genotypes: 48%, 42%, and 36% among all pairs of high peaks for the CRP, Fis, and IHF landscapes, Figure 3d)

Genetic drift facilitates and clonal interference reduces the attainment of high peaks

Taken together, our analyses thus far indicate that high peaks in the CRP, Fis, and IHF landscapes are more accessible than low peaks (Figure 3b), and their basins of attraction also share more genotypes (Figure 3d). However, the accessible evolutionary paths to high peaks are often indirect. In addition, the fraction of accessible paths to any one high peak dwindles rapidly with the distance from that peak (Figure 3a).

To understand how these topographical features jointly affect adaptive evolution, we simulated the evolutionary dynamics on our three landscapes under the assumption that natural selection would favor strong regulation and that fitness is proportional to regulation strength.

Because high peaks constitute only a small fraction of all peaks (2.8%, 0.4%, and 0.5% in the CRP, Fis, and IHF landscapes), we hypothesized that most evolving populations would reach only low peaks. To test this hypothesis, we took advantage of the fact that E.coli has a low genomic mutation rate of 2.2 × 10−10 per base pair per generation 67, and that we study evolution only within an 8-nucleotide segment of a TFBS. In addition, E. coli has a large effective population size (>108 individuals67), which means that genetic drift is weak and even minor differences in fitness are visible to natural selection 67,98. These conditions imply that a population on our landscape would evolve in the well-studied strong-selection weak-mutation regime (SSWM)16,99–101. In this regime, a population is monomorphic most of the time, until a beneficial mutation arises, which usually sweeps rapidly to fixation. In other words, adaptive evolution can be modeled as an adaptive walk, in which each mutational step is taken with a fixation probability that has been derived by Kimura 98. We thus call the resulting adaptive walks Kimura walks (Supplementary Methods 8). We simulated 103 such adaptive walks starting from each of 15,000 randomly and uniformly selected starting (non-peak) TFBS genotypes from each landscape. Each random walk comprised up to 25 mutational fixation steps, unless it reached a fitness peak earlier. Each adaptive walk also accounted for known E. coli mutational biases 102–104 (Supplementary Methods 8).

As we hypothesized, most adaptive walks indeed reach only low peaks (90%, 85%, and 87% percent for the CRP, Fis, and IHF landscapes.). However, the percentage of walks that terminate at high peaks is several-fold greater than the proportion of high peaks. Specifically, 10 percent of walks terminate at high peaks in the CRP landscape, which is 3.6-fold higher than the percentage (2.8%) of high peaks in this landscape. In the Fis and IHF landscape, adaptive walks are 2 and 1.6-fold more likely to terminate at high peaks than expected from the proportion of these peaks among all peaks (7.4% and 8.1% percent) (Figure 4a).

Peak accessibility, contingency and biases in adaptive walks.

a. More than 10% percent of adaptive walks lead to high peaks. The graph shows the cumulative distribution function (CDF, vertical axis) of regulation strength (horizontal axis) attained at the end of 15 million adaptive walks (15,000 starting genotypes × 1,000 adaptive walks each) for each TF landscape (color legend, CRP: green; Fis: orange; IHF: purple). Dashed lines intersect the CDF at points equivalent to the wild-type regulation strength for each TF (color legend). They help to infer the proportion yTF of walks that terminate at high peaks (S>SWT), which is indicated by the numerical values at the graph’s top for each of the landscapes (color legend). b. Paths to high peaks are not much longer than genetic distances from the starting genotype. Colored boxplots summarize, for each of the three landscapes (horizontal axis), the distribution of shortest genetic distances (blue) and actual path lengths (green) for adaptive walks terminating at high peaks. Path lengths are typically less than one mutational step longer than genetic distances. Each box covers the range between the first and third quartiles (IQR). The horizontal line within each box represents the median value, and whiskers span 1.5 times the IQR. Numbers atop each plot indicate means ± one standard deviation. c. Different numbers of high peaks are reached from different starting variants. Violin plots integrated with boxplots show the distribution in the number of high peaks reached (vertical axis) from each starting variant that attained any high peak during 1,000 adaptive walks.

To determine whether high regulatory peaks are more accessible due to chance alone, we compared the empirical landscapes to randomized “null” landscapes. Specifically, we generated these randomized landscapes by permuting regulation strengths across genotypes while preserving the sampled genotype network (Supplementary Methods 7.6). On each randomized landscape, we then performed adaptive-walks (15,000 walks for each of 103 starting genotypes) with the same parameters as for the biological landscape. For all three landscapes, the fraction of adaptive walks reaching high regulatory peaks in the empirical landscapes exceeds that for the randomized landscapes by almost three-fold, a difference that is statistically significant (Supplementary Figure S30, Supplementary Methods 9). In sum, rugged regulatory landscapes strongly constrain evolutionary trajectories, yet do not render the evolution of strong regulation vanishingly rare. Instead, strong regulatory phenotypes remain evolutionarily attainable at levels that exceed null expectations, even though they are reached by only a minority of evolving populations.

The adaptive walks that reached a high peak were short (Figure 4b). On average, they were also no more than half a mutational step longer than the shortest genetic distance between each starting genotype and peak (Figure 4b, CRP landscape: path length 3.2 ± 1.6 [mean±s.d.] vs. genetic distance 2.8 ± 1.2; Fis landscape: path length 3.1 ± 1.5 vs. genetic distance 2.7 ± 1.2; IHF landscape: path length 3.1 ± 1.5 vs. genetic distance 2.7 ± 1.2).

These short paths can be explained by our previous observation that peaks are widely distributed across the genotype spaces of the three landscapes. This distribution makes it easier for populations starting from any (non-peak) genotype to find a local peak through few mutations. Additionally, path lengths realized during adaptive walks tend to be shorter than the average lengths of accessible paths. (Figure 4b).

Because genetic drift can cause evolving populations to escape a low fitness peak and attain a higher fitness peak, we also asked how small population sizes affect the likelihood that a population attains a high fitness peak (Supplementary Material 8, Supplementary Figures S26-28). When simulating adaptive evolution in populations of N=102 individuals, we found that the likelihood that a population reaches a high peak increased by at least 10% percent (to 18%, 20%, and 21% for the CRP, Fis, and IHF landscapes, Supplementary Figures S26-S28).

We also studied how deviations from the strong selection weak mutation regime would affect evolutionary dynamics. In large populations or at high mutation rates, populations tend to be polymorphic and are subject to clonal interference, where the most advantageous of several co-occurring mutations typically dominates and becomes fixed 105–107. To approximate this dynamic, we conducted simulations using “greedy” adaptive walks (Supplementary Methods 8) 25,108. In such an adaptive walk, it is always a genotype’s mutational neighbor with the highest increase in regulation strength that becomes fixed 100,108–110. In other words, a greedy walk starting from a given genotype is deterministic. We, therefore, simulated only one walk for each of the 15,000 randomly chosen (non-peak) starting genotypes. We found that the fraction of greedy walks reaching high regulatory peaks is slightly lower than in the SSWM regime, with 1%, 2%, and 5% fewer walks achieving such peaks in the CRP, Fis, and IHF landscapes, respectively (Supplementary Figure S29). This outcome is expected, because greedy walks prioritize immediate fitness gains at the expense of the ability to discover higher fitness peaks108.

The attainment of peaks is highly contingent on chance events

Because different basins of attraction share many TFBS genotypes (Figure 3d, Supplementary Figures S24-S25), it is possible that adaptive evolution starting from any one genotype can reach different peaks, depending on which mutational paths it takes. In other words, the structure of the landscapes we study may lead to evolutionary contingency 111,112 – the dependence of an outcome of evolution on unpredictable chance events47,111,113. Indeed, non-peak genotypes can access on average around 30% of the total number of high peaks in each landscape (Supplementary Figure S25). To assess the prevalence of evolutionary contingency, we determined how many different peaks are attained by 103 adaptive walks starting from the same randomly chosen genotype. We performed this analysis on 15,000 starting genotypes, focusing on the subset of those starting genotypes from which at least one high peak is reached during the 103 walks. Specifically, 28.6%, 23%, and 24.5% of starting genotypes for the CRP, Fis, and IHF landscapes, reached at least one high peak. Notably, these percentages represent the fraction of starting genotypes that can access at least one high peak, not the total fraction of adaptive walks that lead to a high peak, which is lower (as shown in Figure 4a). For instance, while adaptive walks starting from 28.6% of genotypes reached at least one high peak in the CRP landscape only 10% of all adaptive walks in this landscape end at a high peak (Figure 4a). Importantly, for any one starting genotype where random walks reached at least one high peak, they typically reached more than one high peak (Figure 4c, median±IQR, 4±4, 4±8, 6±6 for the CRP, Fis, and IHF landscapes, respectively). The number of high peaks attained from any one starting genotype was as large as 21, 30, and 33 for the three respective landscapes. Thus, evolutionary contingency is pervasive in all three landscapes. In addition, not only the end point of evolution but also the evolutionary trajectories leading to this end point show contingency. That is, adaptive walks starting from a specific genotype and ending at a specific high peak often take multiple paths to this peak (Supplementary Figure S31).

Biases towards specific evolutionary outcomes have been documented in both empirical and computational studies 18,31,47,111,114–116. They also exist in our landscapes (Figure 4d, Supplementary Figure S31). Figure 4d illustrates both contingency and bias for 103 adaptive walks starting from each of 10 different genotypes in the CRP landscape (thus a total of 104 adaptive walks). Depending on the starting variant, the adaptive walks reached between 1 and 16 high peaks. Whenever different peaks are reached, they are reached by markedly different proportions of adaptive walks. The same holds in the Fis and IHF landscapes (Supplementary Figure S33). Additionally, while multiple routes to each peak can be traversed (Supplementary Figure S31), we observed in all three landscapes biases towards specific paths that are taken more frequently than others (Supplementary Figure S34).

The number of these variants is shown as an integer on top of the panel, and its percentage among all 15,000 starting variants is shown in parentheses (color legend). Underneath it, the panel shows the mean ± 1 s.d. of the number of attained high peaks. a. d. Some high peaks are reached more often than others. We randomly and uniformly sampled 10 starting genotypes from the CRP landscape, started 103 adaptive walks from each, and recorded the number and frequency of distinct high peaks attained in these random walks. Results for each starting genotype are symbolized by a vertical bar. The number of stacks within each bar (delineated by horizontal lines, also indicated by an integer above each bar) indicates the number of high peaks reached by the 103 adaptive walks. Starting variants are ordered in ascending order based on this number of attained peaks. Stack height indicates the fraction of walks that reached the same peak, and is indicated in red, orange, and yellow for the three most frequently attained peaks. See Supplementary Figure S33 for the Fis and IHF landscapes.

Lastly, because sort-seq measurements are subject to experimental uncertainty71,117,118, an important question is whether such noise affects the inferred structure of a regulatory landscape and the resulting evolutionary dynamics. To address this issue we explicitly incorporated empirically estimated measurement uncertainty into our fitness comparisons and repeated pertinent analyses under two conditions: a noise-free landscape (𝒢S) and an “uncertainty-aware” landscape 𝒢S,τ using genotype-specific noise estimates (τ, see Supplementary Figure S35, Supplementary Methods 10). We examined whether noise-induced changes in landscape structure influence evolutionary trajectories. To do so, we compared adaptive walk dynamics between the two kinds of landscapes using identical population genetic parameters. Despite differences in peak counts and local topology, the overall pattern of genotype visitation during adaptive walks was highly similar between the two kinds landscapes. Specifically, the visitation frequency profiles of genotypes were strongly correlated between 𝒢S and 𝒢S,τ landscapes (Spearman’s ρ<0.001 shown in Supplementary Figure S35). This means that genotypes frequently accessed in the noise-free landscape remain frequently accessed in the noisy landscape.

Discussion

We evaluated the ability of three E. coli global regulators to control gene expression through each of more than 30,000 TFBSs for each regulator. To this end, we utilized a synthetic plasmid-based system that facilitates high-throughput fluorescence measurements. This system allowed us to quantify the regulation strength of individual TFBSs by measuring gene expression through GFP fluorescence 74. Additionally, the system insulates the control of transcription from any direct effects library sequences might have on transcription, such as the transcriptional and translational impacts of 5’ untranslated regions on gene expression 119,120. Recent studies have investigated large empirical datasets of cis-regulatory genotypes, examining the impact of sequence variation on gene regulation in both eukaryotes and prokaryotes6,16,72,81,121–124. However, few studies have focused on these interactions from an evolutionary perspective and studied adaptive landscapes of gene regulation, as we do here6,16,123.

We showed that the regulatory landscapes of all three TFs are highly rugged and have multiple peaks. The ruggedness of all three landscapes is also supported by the prevalence of epistasis between pairs of TFBS mutations (Supplementary Table S5). A particularly important form of epistasis is sign epistasis24,93,94, because it can lead to multiple adaptive peaks 24,93,94 (see Supplementary Methods 7.5). Our landscapes contain up to 65% of mutation pairs with sign epistasis, a value that is especially high compared to the almost exclusively additive interactions of mutations in eukaryotic TFs6,125. However, the TFs we study are not exceptions among other prokaryotic TFs. Recent smaller-scale studies have shown that epistatic interactions are common in binding sites of local prokaryotic regulators, such as AraC (sign epistasis in more than 50% of 20 mutants126) and Cl (sign epistasis in 85% of 113 mutants127).

A possible reason for this greater incidence of epistasis lies in the nature of prokaryotic TFBSs. Specifically, prokaryotic TFBSs are at approximately 20bps twice as long as eukaryotic TFBSs80,128 and exhibit symmetries that reflect the dimeric state of their cognate TFs129–131. These factors may increase the likelihood of intramolecular epistasis. Our observations raise important questions for future work, such as why the landscapes of prokaryotic TFBSs differ so dramatically from those of eukaryotic ones. And what do these differences imply for the evolutionary dynamics of gene regulation?

Despite the high ruggedness and pervasive epistasis of these landscapes, we found that a modest fraction of evolving populations can still access the highest regulatory peaks. Specifically, our evolutionary simulations show that 10% of populations with a size typical of E. coli reach one of the highest peaks. This percentage is significantly higher than in randomized landscapes (Supplementary Methods 9; Supplementary Figure S30), which shows that the structure of our regulatory landscapes facilitates access to stronger regulation than expected by chance. We speculate that this property reflects the biological role of global regulators, which coordinate the expression of many target genes11 and thus must operate across a wide spectrum of regulation strengths, while still permitting the evolution of strong regulation when necessary.

The clonal interference that may occur in even larger populations or in populations with high mutation rates reduces the accessibility of high peaks from 10% to 5% of evolving populations (Supplementary Figure S29). Conversely, in small populations where genetic drift can help a population escape from a low peak, this percentage increases to 18%. Numbers like these render the de novo evolution of strong transcription factor binding sites plausible for our three global regulators. Once such binding sites have originated in one population, they can also spread to others through horizontal gene transfer132,133.

In addition to increasing the ruggedness of a landscape, epistasis can also influence the evolution of populations on the landscape24,94,134. The regulation strength of a TFBS is partially determined by the combined interactions of the nucleotides within the TFBS50,135–137, which in turn affects TF-TFBS binding affinities. Different combinations of mutations can result in similar regulation strengths and create multiple evolutionary pathways to achieve optimal or near-optimal regulation6,138 (Supplementary Figure S31). This can lead to overlapping basins of attraction among different peak TFBSs conveying strong regulation, such that even evolution starting from the same genotypes can take different paths and reach different peaks with similar regulation strengths. Indeed, we observed such overlapping basins in our landscapes (Figure 3d and Supplementary Figure S24). Moreover, epistasis can create plateaus of regulation strength where various combinations of nucleotides yield TFBSs with intermediate regulation strengths. These plateaus further contribute to the overlapping basins of attraction we observe, and may serve as common evolutionary intermediates that multiple starting sequences can traverse on their way to different peaks.

In a landscape with substantial epistasis, the sequence of mutations that occurs in an evolving population can also render adaptive evolution highly contingent on this sequence111,112. For example, some beneficial mutations within a TFBS might only be accessible after other specific mutations have occurred. Different starting sequences or early mutations can also change the spectrum of accessible mutations, leading to different peaks with similarly strong regulation 111,112. Indeed, we observed in all three landscapes that different evolving populations starting from the same genotypes in the landscape attain different peaks (Figures 3d, 4d and Supplementary Figure S33) 111,113,139,140. Such contingency reduces the predictability of evolution47,111,141.

Despite the prevalence of contingency in all three landscapes, evolving populations starting from the same genotype more often attain some peaks than others (Supplementary Figure S33). Moreover, for each attained peak with multiple possible evolutionary routes, we also observed that some paths are more frequently transversed than their alternatives (Supplementary Figure S34). That is, evolution is biased towards traversing some paths and attaining some peaks more often than others142,143. These observations emphasize the complex interplay between chance, contingency, and evolutionary biases in shaping the outcomes of adaptive evolution.47,111,113,115.

The three TFs we studied here belong to different protein families, yet they have regulatory landscapes with similar topography. All three landscapes are highly rugged, highly epistatic, and harbor multiple peaks that are widely scattered, and this holds for peaks of low, intermediate, and high regulation strength. In addition, peak accessibility in all three landscapes increases with peak height (Supplementary Figure S32). On the one hand, these commonalities may be caused by common biological or biochemical properties of the three TFs. For example, CRP has been suggested to possess nucleoid-associated protein properties similar to Fis and IHF, due to its ability to bend and loop DNA144. On the other hand, the commonalities may reflect general characteristics of global gene regulation. One of them is that global TFs often bind unspecifically to multiple TFBSs11,13. Also, they may bind DNA with a broad range of different affinities centered around intermediate affinity, rather than bind few sites but very strongly 13. This possibility is supported by comprehensive analyses of in vitro eukaryotic TF binding affinities for thousands of TFBS variants, showing essentially continuous binding affinity distributions for hundreds of sites6.

Despite broad similarities among the three landscapes we study, we also found some differences. Most notable is the narrower distribution of regulation strengths for the binding sites of CRP compared to those of Fis and IHF (Figure 2a). This is not unexpected, given that CRP binding sites are known for being quasi-symmetric and less degenerate than those of Fis and IHF57,60. Previous studies have also shown that Fis and IHF binding sites are less conserved in their nucleotide sequence and more biased toward AT-rich content43,145,146. Differences in the distribution of regulation strengths may also result from differences in functions among the three proteins. CRP’s function is mostly restricted to gene regulation. As a result, it may have evolved to finely tune its regulatory output, resulting in a narrower distribution of regulation strengths. In contrast, Fis and IHF are involved in functions beyond gene regulation, such as DNA replication and genomic structural maintenance147,148, which might impose additional constraints or require a broader range of DNA binding strengths.

One limitation of our work stems from our rigorous quality filtering of sort-seq data. As a result, we lack regulatory data for some 40% of the TFBSs in each landscape. This limited diversity of reliable data is a common feature in mutational library studies. It has several technical causes, such as biases in library synthesis 149, PCR amplification 150, cloning, and loss of sequence diversity after cell sorting 122. In future work, this limitation could be overcome by a combination of strategies, such as subsampling complete genotype spaces or combining different molecular methods to overcome biases and diversity loss during PCR amplification151,152, sorting71,117,153, and high-throughput sequencing123.

Importantly, although undersampling of genotype space is a limitation, standard approaches such as random subsampling or predictive modeling are not straightforward remedies. Several of our core analyses—including peak identification, quantification of epistasis, and assessment of evolutionary accessibility—rely on combinatorially complete local neighborhoods in genotype space. Random subsampling of genotypes would remove mutational neighbors, and thereby confound pairwise comparisons and the interpretation of landscape topology. Predictive modeling could be used to infer missing genotypes and reconstruct more complete landscapes154, but it requires additional assumptions that introduce their own limitations. In addition, developing, validating, and benchmarking such models would be beyond the scope of this study, which is focused on empirical landscape mapping, and can serve as a starting point for future modeling work.

A second limitation comes from our use of the sort-seq method, which is best-suited for our work, because it allows the high-throughput measurement and sorting of millions of individual cells in a standardized and straightforward manner. However, the method’s accuracy depends on the binning procedure. Other studies have used between two and several dozen bins, with or without unbiased sampling 71,73,81,117,121,122,155. Recommendations emerging from this work include sorting of cells into at least four logarithmically (log2) equally-spaced bins, each covering approximately 12.5-15% of the fluorescence distribution, 81,122,156. We followed these recommendations, using 13 bins. In addition, we computed a weighted average to calculate expression values from sequences appearing in multiple bins, a straightforward method validated by robust studies in transcriptional regulation analysis 16,72. In addition, we validated individual regulation strengths with an independent method to demonstrate its reliability.

A third limitation of our study is that we only examined variation at eight positions for each TF. We selected these positions based on their importance for DNA binding and regulation by our TFs, as evidenced by their high information content. We cannot exclude the possibility that selecting a different set of positions could yield different landscape topographies. However, we speculate that less information-rich position would not reduce but rather increase the potential for the de novo evolution of strong TFBSs. For example, they may facilitate landscape navigability through extradimensional bypasses96,157 or provide small-incremental changes in regulation strengths. These small changes help populations reach peaks via diminishing returns effects, meaning that as a population gets closer to a peak, each subsequent mutation contributes progressively smaller improvements to regulation strength 158,159. Investigating a wider array of positions and larger regulatory landscapes remains an important task for future work.

Fourth, we use a simplified empirical system for the study of gene regulation—a promoter followed by a single TFBS. Although this regulatory architecture exists for some genes160, global regulators often form part of more complex, combinatorial architectures4,160,161 that are influenced by environmental factors2,162,163, concentrations of active TFs164–166, chromosome structure167–171, methylation states172, and more. Studies on more complex promoter architectures may reveal regulatory landscapes with different topographies.

Lastly, we also analyzed the adaptive evolution of TFBSs in a simplified manner, performing data-driven evolutionary simulations rather than experimental evolution. Although many studies assume that regulation strength and fitness are correlated, this is not always the case. Low binding affinities can be adaptive during development 173, and many genes exhibit a nonlinear fitness-expression function 174 with a plateau of maximal fitness across a wide range of expression levels 175–177. For instance, most mutations and polymorphisms in the promoter of the yeast gene TDH3 do not significantly affect fitness in a glucose-rich medium 176. For these reasons we focused our observations and interpretations on regulatory phenotypes (regulation strength). Our simplified approach allowed us to model population evolution in a large genotype space and avoid monitoring thousands of evolving regulatory sequences simultaneously in vivo23,178. However, it cannot replace experimental evolution of TFBSs, which also remains an important challenge for future work.

Transcription factor binding sites are among the simplest units of biological organization. Our work provides the first large-scale analysis of the regulatory landscapes formed by such sites for global transcriptional regulators. It shows that strong binding sites can readily evolve de novo, even though prokaryotic transcription factor binding sites are much larger than their eukaryotic counterparts and have more rugged regulatory landscapes. In addition, the evolution of these simple sequences also displays phenomena that have been characterized in much more complex systems, such as evolutionary contingency and evolutionary biases.

Materials and methods

The Supplementary Materials contain extended details of experimental procedures and data analysis.

Strains and plasmids

Bacterial strains and plasmids used in this work are listed in Table S1. We obtained electrocompetent E. coli cells of strain SIG10-MAX® from Sigma Aldrich (CMC0004). We used this strain for molecular cloning and library generation due to its high transformation efficiency. The genotype of this strain (Supplementary Table S4) is similar to DH5α (Sigma Aldrich commercial information, see Supplementary Table S4). The strain is resistant to the antibiotic streptomycin.

We amplified plasmid libraries in SIG10-MAX®, extracted their DNA, and transformed them into mutants derived from E. coli K-12 strain BW25113 that harbor chromosomal deletions of the crp, fis, or ihfa gene. We obtained these mutant strains from the KEIO collection 179 and used them for sort-seq experiments. The design, genetic parts, and assembly of the plasmid vectors we used in this study are available in the Supplementary material, as are all primers, TFBS sequences/libraries, strains, and plasmids.

Sort-Seq procedure

To explore the regulatory effects of each transcription factor on binding sites in the corresponding library, we constructed three plasmids, each of which enables the inducible expression of one of our three TFs. These are plasmids pCAW-Sort-Seq-V2-CRP, pCAW-Sort-Seq-V2-Fis, and pCAW-Sort-Seq-V2-IHF (Supplementary Methods 3 and Supplementary Table S1). We cloned the TFBS libraries (Supplementary Table S3) into their respective plasmids and then transformed them into mutant strains lacking the corresponding TFs (Δcrp, Δfis, and Δihf, as listed in Supplementary Table S2). We induced TF expression using anhydrotetracycline (Atc) and, after overnight growth, performed cell sorting for cell populations. During sorting, we distributed cells into 13 equally-spaced logarithmic bins based on their fluorescence levels. We replicated each sort-seq experiment three times from three separate library transformations for each TF. To mitigate the impact of extrinsic noise (gene expression variation among cells 180), we adhered to standard protocols by normalizing our GFP fluorescence measurements against mScarlet-I fluorescence values obtained from flow-cytometry assays 7,71,181,182. We subsequently recovered the sorted cells from each bin in 50mL Falcon tubes containing 10 mL of LB medium supplemented with chloramphenicol and incubated them overnight at 37°C with shaking at 220 rpm. After this growth period, we re-sorted cell cultures from each bin to eliminate potential contaminants and ensure that the cell populations had preserved their fluorescence distributions. Following re-sorting, we extracted plasmids from the cell population of each bin. We amplified and barcoded the TFBS region from each population through a polymerase chain reaction (PCR). Lastly, we sequenced barcoded amplicons containing TFBS sequences, and used the sequencing results to calculate the regulation strengths of each TF to its TFBSs in the corresponding library. More details are provided in the Supplementary Material.

Regulation strengths

Due to gene expression and measurement noise, individual TFBS variants in a sort-seq experiment usually appear in more than a single bin, and their read count (frequency) varies among bins71,72,121,153. Following established practice18,72,123, we used a weighted average of these frequencies for each variant to represent the mean expression level caused by the variant. To facilitate the interpretation of this quantity, we converted this expression level into a regulation strength relative to the highest observed regulation strength for a given TF, to which we assigned a value of one. From each library, we selected a single naturally occurring binding site for each TF that was previously characterized in the literature as a strong binder. We called this TFBS the WT sequence and used it as a baseline to separate weakly (higher GFP expression) from strongly (lower GFP expression) regulating TFBSs.

Validating regulation strengths with plate reader measurements

To further validate our regulation strength data, we chose 10 DNA binding sites from each bin (10 variants ×13 bins = 130 variants in total per TF library, Supplementary File 2), covering a wide range of measured regulation strengths. We cloned these sequences into the appropriate vector pCAW-Sort-Seq-V2-TF (TF: CRP, Fis or IHF), and transformed them into the appropriate mutant strain. We picked individual colonies and grew them overnight (16 hours, 37°C, 220 rpm) in liquid LB supplemented with 50 μg/mL of chloramphenicol and anhydrotetracycline. We diluted the cultures to 1:10 (v/v) in cold Dulbecco’s PBS (Sigma-Aldrich #D8537) to a final volume of 1 mL. We transferred 200 μl of the diluted cultures into individual wells in 96-well plates and measured GFP fluorescence (emission: 485nm/excitation: 510nm, bandpass: 20nm, gain: 50), as well as the optical density at 600nm (OD600) as an indicator of cell density. We then normalized fluorescence by the measured OD600 value to account for differences in cell density among cultures and compared the obtained ratios to the previously inferred regulation strengths for the 130 selected variants. We performed all such measurements in biological and technical triplicates (three colonies per sample, and three wells per colony, respectively).